### 2016-08-29 ms/notes/occupancy/notes.tex
added manuscript folder

\documentclass[12pt]{article}
\usepackage[top=0.85in, bottom=0.93in, right=0.93in, left=0.93in,
paperwidth=8.5in, paperheight=11in, nohead]{geometry}
\geometry{letterpaper}
\usepackage[pdftex]{graphicx}
\usepackage{color}
\usepackage[normalem]{ulem}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{setspace}
\usepackage{mdwlist}
\usepackage{verbatim}
\usepackage{tikz}
\usepackage[numbers]{natbib}

%% make bibliography more compact \setlength{\bibsep}{0in}
\renewcommand\bibsection{\subsubsection*{\refname}}

\begin{document}

\begin{centering}
  \large {\bf Occupancy models with NIMBLE} \\
\end{centering}
\vspace{0.15in}

\section{Goals}
\begin{enumerate}
\item Showcase the functionality/efficiency of NIMBLE using occupancy
  modeling
\item provide code online and as supplement for common occupancy
  models (and a tutorial?)
\item target journal: \textit{Methods in Ecology and Evolution}
\end{enumerate}

\section{Models}
\begin{enumerate}
\item Multi-season occupancy model
  \begin{enumerate}
  \item vectorize Bernoulli call \checkmark
  \item write a user-defined MCMC for the Zs, slice sampler for other
    parameters \checkmark
  \item custom z sampelr and auto blocking for other paramters
    \checkmark
  \item add reflective samplers for bounded parameters, and custom z
    sampler \checkmark
  \item write function for latent state (Zs) likelihood \checkmark
  \item write function for latent state (Zs) likelihood with auto
    blocking \checkmark
  \item sample uniforms on logit scale?
  \item cross level sampling?
  \end{enumerate}
\item Multi-species occupancy model (from Zipkin et al.)
  \begin{enumerate}
  \item write function for latent state (Zs) likelihood \checkmark
  \item block samplers for species random effect for each type \checkmark
  \item block samplers for species random effect for each species
    across types \checkmark
  \item for all above, compute ``derived quantities'' as posterior
    sample of probabilities vs. posterior sample of binomial samples
    of the posterior probabilities \checkmark (now removed from
    efficiency calculation)
  \end{enumerate}
\item Spatial occupancy model (CAR model or distance decay)
  \begin{enumerate}
  \item have to implement the model
  \item auto-blocking of parameters related to decay with distance
  \end{enumerate}
\end{enumerate}

\section{Comparisons}
\begin{enumerate}
\item original single species, multi season model to 1) vectorized
  Bernoulli draw, 2) custom z sampler with slice sampler on other
  parameters, 3) custom z sampler with auto blocker on other
  paramters, 4) custom z sampler with reflective sampler on other
  parameters, 5) explicit likelihood for latent state, 6) explicit
  likelihood for latent state and auto blocking
\item original multi species, single season model to 1) explicit
  likelihood for latent state with vectorized calculations and
  expected values of derived quantities, 2) all the same but with
  block sampling of the species random effects
\end{enumerate}

\section{Results}
\subsection{Single species, multi season}
\begin{enumerate}
\item JAGS and NIMBLE tie on the multi season model when the slice
  sampler and custom z sampler are used (same samplers as JAGS)
\item For the run with custom z on the zs and slice samplers on the
  uniform(0,1)s, copying takes about $10\%$ of total time.
\item For the run with custom z on the zs and default RW samplers on
  the uniform(0,1)s, copying takes a bit over $20\%$ of total time.
\item So not trivial, but also not something that if improved would
  cause a huge jump in performance.  Improvements from algorithmic
  insights have potential for big jumps.
\end{enumerate}

\section{Considerations}
\begin{enumerate}
\item Daniel just modified MCMCsuite() to allow a skip check argument
  to allow user-defined functions with dimension ``mismatches'' to
  pass the check \checkmark
\item try MCMCsuite() with user function with and without check
\item compareMCMC() is a new function, see Perry tutorial for an
  example (make sure summary=FALSE to allow for further comparisons)
\end{enumerate}

\section{Misc. notes}
\begin{enumerate}
\item copy issue in nimble:  In Jags each step always updates, whereas
  in NIMBLE certain proposals can be rejected. This enables NIMBLE to
  be more flexible about MCMC samplers (for example, Metropolis
  Hastings) but means NIMBLE has to copy more. If the proposal is
  rejected the original is copied to the next step. If the proposal is
  accepted, the new value because the original for the next step.
\item thinning may ''hurt'' samples that are mixing well. In contrast,
  a poorly mixing sampler will be ``helped'' because tinning helps is
  ``mix better''
\item reflective samplers and bounded parameters: Reflective samplers
  might be most helpful when the posterior is skewed. In that case,
  the sampler would reject more often near the boundaries in the
  non-reflective case, so it may take longer to converge on the
  correct, skewed posterior. In constrast, when the posterior is not
  skewed, it may not matter as much because the sampler will spend
  less time at the doundaries.
\end{enumerate}

\section{Future directions}
\begin{enumerate}
\item ``parametric bootstrapping'' for model evaluation, calibrated
  pvalues
\item avoiding the data augmentation BUGS work around paper
\end{enumerate}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-PDF-mode: t
%%% End:

### 2016-08-29 singleSpp-multiSea/compare.R
removed old figures

@@ -42,8 +42,8 @@ ss.ms.occ.all <- combine_MCMC_comparison_results(ss.ms.orig[[1]],
                                                  ss.ms.opt1[[1]],
                                                  ss.ms.opt2[[1]],
                                                  ss.ms.opt3[[1]],
-                                                 ss.ms.opt4[[1]]
-                                                 ss.ms.opt5[[1]]
+                                                 ss.ms.opt4[[1]],
+                                                 ss.ms.opt5[[1]],
                                                  name = "ss.ms" )
 
 make_MCMC_comparison_pages(ss.ms.occ.all, dir="figures/comparisons")

### 2016-09-01 ms/ms/ms.tex
finished multip species runs

\documentclass[12pt]{article} 
\usepackage[pdftex]{graphicx}
\usepackage{natbib} 
\usepackage{color}
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{verbatim}
\usepackage{mathpazo} 
\usepackage{setspace}
\usepackage{multirow}
\usepackage{fullpage}
\usepackage{lscape}
\usepackage{fancyhdr}
\usepackage[normalem]{ulem} 
\usepackage{hyperref}
\usepackage[parfill]{parskip}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xr}

\hypersetup{colorlinks=true, linkcolor=black, citecolor=black}
\RequirePackage{lineno}

\newcommand{\flagged}[1] {
  \textcolor{blue}{#1}
}

\def\title{Increasing the tractability of occupancy models}
\def\author{Lauren C.\ Ponisio$^{1,2}$, Perry de Valpine$^1$}

\def\runninghead{Occupancy model efficiency}
\def\keywords{NIMBLE, Markov chain Monte Carlo, latent states, block
  sampling, dynamic occupancy, mutli species occupancy, spatial
  occupancy, JAGS }

\def\extras{
  \begin{itemize}
  \item Submitted as a Standard paper
  \item Abstract word count: 
  \item Main text word count: 
  \item Number of references: 
  \item Number of figures:
  \end{itemize}
}

\def\affiliation{
  \begin{enumerate}
  \item Department of Environmental Science, Policy, and Management\\
    University of California, Berkeley\\
    130 Mulford Hall\\
    Berkeley, California, USA\\
    94720\\
  \item Department of Entomology\\
    University of California, Riverside\\
    417 Entomology Bldg.\\
    Riverside, California, USA\\
    92521\\
  \end{enumerate}
}

\newcommand{\mstitlepage}{
  \paragraph{Running head:} \textsc{\runninghead}
  % \parindent=0pt
  \begin{center}%
    {\LARGE \title \par}%
    \vskip 3em%
    {\large
      \lineskip .75em%
      \begin{tabular}[t]{c}%
        \author
      \end{tabular}\par}%
    \vskip 1.5em%
  \end{center}\par
  \affiliation
}
\clearpage

\begin{document}

\mstitlepage
\doublespacing
\linenumbers
\clearpage

\begin{abstract}  
place holder
\end{abstract}

\keywords

\clearpage

\section*{Introduction}
\label{sec:introduction}

Estimating the proportion of sites occupied by a species is common
challange for many subdisciplines ecology and evolution including
metapopulation, engangered and invasion biology. Greater acceptance of
the biases of imperfect detection has lead to the development and
proliferation of occupancy models, which model the occurrence of a
species at a site as a latent state layered underneath a detection
process \citep[e.g.,][]{mackenzie-2006, royle-2007-1813}. Now only a
little over a decade after occupancy models were introduced to
ecology, they are being used to model the occurence of everything from
bees \citep{mgonigle-2015-x} to tigers \citep{hines2010tigers} in an
endless variety of complexity.

Occupancy models are part of a larger class of models known as Hidden
Markov Models. For discrete Hidden Markov Models like occupancy models
where a species is either present or absent from a site, likelihood
calculation involves summing over the distribution of latent
states. Because estimating the effect of explanatory variables on site
occupancy or shared variation of in occupancy across species is often
of greatest interest to ecologsits
\citep[e.g.,][]{iknayan2014detecting}, the Hidden Markov Models are
embedded in a larger hierarchical model. In such cases, practitioners
may rely on Markov chain Monte Carlo (MCMC) to perform a Bayesian
analysis. Such models are computationally intensive, and large models
requiring hundreds or thousands of dimensions which require MCMC can
be computationally impractical. To ensure occupancy models are
accessible to practitioners, more efficient methods for estimating these
models are necessary.

%% model selection

\section*{Materials \& Methods}
\label{sec:methods}

\section*{Results}
\label{sec:results}

\section*{Discussion}
\label{sec:discussion}

\section*{Acknowledgments}
\label{sec:acknowledge}

\bibliographystyle{mee}

\bibliography{refs}

% \begin{figure}
%   \centering
%   \includegraphics[width=.8\textwidth]{../analysis/changePoint/plotting/networks.pdf}
%   \caption{Assembling hedgerow networks had more changing points
%     (vertical red lines) than non-assembling hedgerows and weedy field
%     margins (a representative sample of non-assembling sites are
%     depicted here). In each network, plants and pollinators are
%     represented by green and yellow circles, respectively, weighted by
%     their degree. Each species has a consistent position in the
%     network across years. In the assembling hedgerows, colored squares
%     in the corner of each network represent the year of
%     assembly. Before hedgerows are installed (when they are still
%     field margins) the year of assembly is zero.}
%   \label{fig:changePoints}
% \end{figure}
% \clearpage

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-PDF-mode: t
%%% End:

### 2016-09-01 multiSpp-singleSea/compare.R

@@ -14,9 +14,9 @@ load(file="saved/opt1.Rdata")
 ## species
 load(file="saved/opt2.Rdata")
 
-## option 1 + custom block sampler on species random effect for each
-## random effect type
-load(file="saved/opt3.Rdata")
+## ## option 1 + custom block sampler on species random effect for each
+## ## random effect type
+## load(file="saved/opt3.Rdata")
 
 ## option 1 + sigma sampler on random effects
 load(file="saved/opt4.Rdata")
@@ -25,9 +25,9 @@ ms.ss.opt1[[1]] <- rename_MCMC_comparison_method('nimble', 'remove_z',
                                                  comparison=ms.ss.opt1[[1]])
 ms.ss.opt2[[1]] <- rename_MCMC_comparison_method('nimbleOpt2', 'block_1',
                                                  comparison=ms.ss.opt2[[1]])
-ms.ss.opt3[[1]] <- rename_MCMC_comparison_method('nimbleOpt3',
-                                                 'block_2',
-                                                 comparison=ms.ss.opt3[[1]])
+## ms.ss.opt3[[1]] <- rename_MCMC_comparison_method('nimbleOpt3',
+##                                                  'block_2',
+##                                                  comparison=ms.ss.opt3[[1]])
 ms.ss.opt4[[1]] <- rename_MCMC_comparison_method('nimbleOpt4',
                                                  'sigma sampler',
                                                  comparison=ms.ss.opt4[[1]])
@@ -36,7 +36,7 @@ ms.ss.opt4[[1]] <- rename_MCMC_comparison_method('nimbleOpt4',
 ms.ss.occ.all <- combine_MCMC_comparison_results(ms.ss.orig[[1]],
                                                  ms.ss.opt1[[1]],
                                                  ms.ss.opt2[[1]],
-                                                 ms.ss.opt3[[1]],
+                                                 ## ms.ss.opt3[[1]],
                                                  ms.ss.opt4[[1]],
                                                  name = "ms.ss" )
 

### 2016-09-02 ms/ms/ms.tex
added abstract to ms

@@
 \def\title{Increasing the tractability of occupancy models}
-\def\author{Lauren C.\ Ponisio$^{1,2}$, Perry de Valpine$^1$}
+\def\author{Lauren C.\ Ponisio$^{1,2}$, Nicholas Michaud$^1$, Perry de
+  Valpine$^1$}
 
@@
 \begin{abstract}  
-place holder
+  \begin{enumerate}
+  \item occupancy models are everywhere, but model fitting and
+    assessment are extrememly computationally intensive
+  \item Using the NIMBLE package for R, we develop combined
+    computational approaches to improve computational efficiency
+    including user defined and automatic blocking of parameters for
+    MCMC, filtering over latent states, and customized MCMC samplers
+    for specific parameters. We test these approaches for three
+    representative occupancy models of varying levels of complexity
+    including a single species dynamic (i.e., multi-season) model with
+    spatial auto correlation, and a mutli-species model. We also use
+    calibrated posterior $p$-values to assess model fit.
+  \item These computation approaches lead to an improvement in MCMC
+    sampling efficiency, particularly with the more complex,
+    mutli-species model.
+  \item Ours results highlight the need for more customizable
+    approaches to MCMC to fit and assess hierachical models in order
+    to ensure occupancy models are accessible to practitioners. By
+    implementing MCMC procedures and model assessment techniques open
+    source software, we have made progress toward tis aim.
+  \item \textit{Implications:}
+  \end{enumerate}
 \end{abstract}
@@
 requiring hundreds or thousands of dimensions which require MCMC can
-be computationally impractical. To ensure occupancy models are
-accessible to practitioners, more efficient methods for estimating these
-models are necessary.
-
-%% model selection
+be computationally intensive. In addition, fitting these models is
+such a challenge that users often forgo adding any additional
+computation to asses model fit. To ensure occupancy models are
+accessible to practitioners, more efficient methods for fitting and
+assessing these models are necessary.
 

### 2016-09-02 spatial/src/plotting.R

@@ -1,35 +1,0 @@
-
-  pdf.f <- function(f, file, ...) {
-    cat(sprintf('Writing %s\n', file))
-    pdf(file, ...)
-    on.exit(dev.off())
-    f()
-  }
-
-## function to plot values of parameters as a function of
-## iterations. Input is a MCMCcompare object
-
-checkChains <- function(all.mods.samps, f.path){
-  niter <- dim(samp.all)[3]
-  params <- names(samp.all[1,,1])
-  mods <- names(samp.all[,1,1])
-  if(is.null(mods)) mods <- "nimble"
-
-  lapply(mods, function(z){
-    f <- function(){
-      layout(matrix(1:4, ncol=2))
-
-      apply(samps[z,,], 1, function(x){
-        plot(x, type="l",
-             xlab = 'iteration',
-             main= params[which(apply(samps[z,,], 1, function(y)
-               all(match(x,y))))]
-             )
-      })
-    }
-
-    pdf.f(f,
-          file= file.path(sprintf(f.path, z)),
-          height=11, width=8.5)
-  })
-}

### 2016-09-02 ms/ms/ms.tex
cppp function working

@@
     MCMC, filtering over latent states, and customized MCMC samplers
-    for specific parameters. We test these approaches for three
+    for specific parameters. We test these approaches using three
     representative occupancy models of varying levels of complexity
@@
     implementing MCMC procedures and model assessment techniques open
-    source software, we have made progress toward tis aim.
+    source software, we have made progress toward this aim.
   \item \textit{Implications:}

### 2016-09-02 cppp/src/calcCPPP.R

@@ -1,0 +1,95 @@
+library(parallel)
+## draws a random number from the posterior, reassigned it as the
+## paramter value, simulates data from the model
+pppFunc <- nimbleFunction(
+  setup = function(model, dataName, paramNames, MCMCmv, MCMCIter, thin){
+    paramDependencies <- model$getDependencies(paramNames)
+  },
+  run = function(N = integer(0)){
+    deviances <- numeric(N)
+    for(i in 1:N){
+      randNum <- ceiling(runif(1, 0, MCMCIter/thin -1 ))
+      nimCopy(MCMCmv, model, paramNames, row = randNum)
+      calculate(model, paramDependencies)
+      simulate(model, dataName)
+      deviances[i] <- calculate(model)
+    }
+    return(deviances)
+    returnType(double(1))
+  }
+  )
+
+
+
+generateCPPP <-  function(R.model,
+                          C.model,
+                          C.mcmc,
+                          mcmc,
+                          dataName,
+                          paramNames, ## vector of parameters to monitor
+                          MCMCIter, ## number of samples
+                          NSamp,
+                          NPDist,
+                          thin,...){
+  ## calculate proportion of deviances that are greater or equal to
+  ## the observed value
+  calcCPPP <- function(MCMCIter, C.model, NSamp,
+                       C.pppFunc, C.mcmc){
+
+    C.mcmc$run(MCMCIter)  
+    observedDisc <- calculate(C.model)  
+    otherDiscs <- C.pppFunc$run(NSamp)
+    pre.pp <- mean(otherDiscs >= observedDisc)
+    return(pre.pp)    
+  }
+
+  ## simulate the distribution of posterior predictive pvalues
+  simPppDist <- function(interation,
+                         MCMCIter,
+                         C.mcmc,
+                         C.model,
+                         paramNames,
+                         paramDependencies,
+                         NSamp, thin){
+    randNum <- ceiling(runif(1, 0, MCMCIter/thin - 1))
+    mcmc.samples <- as.matrix(C.mcmc$mvSamples)
+    values(C.model, paramNames) <- mcmc.samples[randNum,]
+    calculate(C.model, paramDependencies)
+    simulate(C.model, dataName)
+    out <- calcCPPP(MCMCIter, C.model, NSamp, C.pppFunc, C.mcmc)
+    return(out)
+  }
+
+  ## sample posterior, simulate data from sample 
+  paramDependencies <- C.model$getDependencies(paramNames)
+  mcmcMV <- mcmc$mvSamples
[…]

### 2016-09-02 cppp/testcppp.R

@@ -1,26 +1,9 @@
+rm(list=ls())
 library(nimble)
 library(parallel)
 options(mc.cores=2)
-
-pppFunc <- nimbleFunction(
-  setup = function(model, dataName, paramNames, MCMCmv, MCMCIter){
-    postLength <- MCMCIter
-    paramDependencies <- model$getDependencies(paramNames)
-  },
-  run = function(N = integer(0)){
-    deviances <- numeric(N)
-    for(i in 1:N){
-      randNum <- ceiling(runif(1, 0, postLength))
-      nimCopy(MCMCmv, model, paramNames, row = randNum)
-      calculate(model, paramDependencies)
-      simulate(model, dataName)
-      deviances[i] <- calculate(model)
-    }
-    return(deviances)
-    returnType(double(1))
-  }
-  )
-
+nthin <- 2
+source('~/Dropbox/occupancy-nimble/cppp/src/calcCPPP.R')
 
 
 pumpCode <- nimbleCode({
@@ -40,93 +23,36 @@ pumpData <- list(x = c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22))
 pumpInits <- list(alpha = 1, beta = 1,
                   theta = rep(0.1, pumpConsts$N))
 
-generatecPPP <-  function(code, ## model created by nimbleCode()
-                          constants,
-                          data,
-                          dataName,
-                          inits, ## inital values for parameters to estimate
-                          paramNames, ## vector of parameters to monitor
-                          MCMCIter, ## number of samples
-                          NSamp,
-                          NPDist,
-                          thin,...){ ## thinning rate
-  calcCPPP <- function(MCMCIter, C.model, NSamp){
-    C.mcmc$run(MCMCIter)  
-    observedDisc <- calculate(C.model)  
-    otherDiscs <- C.pppFunc$run(NSamp)
-    pre.pp <- mean(otherDiscs >= observedDisc)
-    return(pre.pp)    
-  }
 
-  ## build model
-  R.model <- nimbleModel(code=code,
-                         constants=constants,
-                         data=data,
-                         inits=inits,
-                         check=FALSE,...)
-  message('R model created')
-  
-  ## configure and build mcmc
-  mcmc.spec <- configureMCMC(R.model,
-                             print=FALSE,
-                             monitors = paramNames,
-                             thin=thin)
-  mcmc <- buildMCMC(mcmc.spec)
-  message('MCMC built')
-  
[…]

### 2016-09-02 multiSpp-singleSea/src/initialize.R

@@ -1,6 +1,6 @@
 ## library(devtools)
 ## install_github("nimble-dev/nimble",
-##                ref = "newNodeFxns",
+##                ref = "devel",
 ##                subdir = "packages/nimble")
 
 library(nimble)

### 2016-09-02 multiSpp-singleSea/src/multispeciesOcc.R

@@ -4,32 +4,34 @@
 
 dBernDetectionMatrix <- nimbleFunction(
   run = function(x = double(2),
-                 occProb = double(1),
-                 detectionProb = double(2),
-                 num_reps = double(1),
-                 log = double(0, default = 0)) {
+    occProb = double(1),
+    detectionProb = double(2),
+    num_reps = double(1),
+    log = double(0, default = 0)) {
     returnType(double())
     num_points <- dim(x)[1]
     ans <- 0
     ## new code 5/25/16
     for(j in 1:num_points) {
-        probDetectionHistoryGivenOccupied <- 1
-        probDetectionHistoryGivenUnoccupied <- 1
-        for(k in 1:num_reps[j]) {
-            if(x[j,k] == 1) {
-                probDetectionHistoryGivenOccupied <- probDetectionHistoryGivenOccupied * detectionProb[j,k]
-                probDetectionHistoryGivenUnoccupied <- 0
-            } else {
-                probDetectionHistoryGivenOccupied <- probDetectionHistoryGivenOccupied * (1-detectionProb[j,k])
-            }
+      probDetectionHistoryGivenOccupied <- 1
+      probDetectionHistoryGivenUnoccupied <- 1
+      for(k in 1:num_reps[j]) {
+        if(x[j,k] == 1) {
+          probDetectionHistoryGivenOccupied <-
+            probDetectionHistoryGivenOccupied * detectionProb[j,k]
+          probDetectionHistoryGivenUnoccupied <- 0
+        } else {
+          probDetectionHistoryGivenOccupied <-
+            probDetectionHistoryGivenOccupied * (1-detectionProb[j,k])
         }
-        ans <- ans + log(occProb[j] * probDetectionHistoryGivenOccupied + (1-occProb[j]) * probDetectionHistoryGivenUnoccupied)
- ## old code that looks like it was wrong:       
- ##       for(k in 1:num_reps[j]) {
- ##       if(x[j,k] == 1) ans <- ans + log(occProb[j] * detectionProb[j,k])
- ##       else ans <- ans + log(occProb[j] * (1-detectionProb[j,k]) +
- ##                             (1-occProb[j]))
- ##     }
+      }
+      ans <- ans + log(occProb[j] * probDetectionHistoryGivenOccupied + (1-occProb[j]) * probDetectionHistoryGivenUnoccupied)
+      ## old code that looks like it was wrong:       
+      ##       for(k in 1:num_reps[j]) {
[…]

### 2016-09-02 singleSpp-multiSea/opt4-5.R

@@ -51,6 +51,7 @@ input1 <- list(code=ss.ms.occ,
                data=model.data,
                inits=inits)
 
+monitors <- c("phi", "gamma", "psi")
 
 ss.ms.opt4 <- compareMCMCs(input1,
                            MCMCs=c('nimble', 'autoBlock'),
@@ -85,6 +86,7 @@ MCMCdefs.opt5 <- list('nimbleOpt5' = quote({
 
 ## *********************************************************************
 ## run with compareMCMCs
+## *********************************************************************
 
 input1 <- list(code=ss.ms.occ,
                constants=constants,
@@ -101,3 +103,48 @@ ss.ms.opt5 <- compareMCMCs(input1,
                            check=FALSE)
 
 save(ss.ms.opt5, file="saved/opt5.Rdata")
+
+
+## *********************************************************************
+
+## build model
+R.model <- nimbleModel(code=ss.ms.occ,
+                       constants=constants,
+                       data=model.data,
+                       inits=inits,
+                       check=FALSE)
+message('R model created')
+
+
+## configure and build mcmc
+mcmc.spec <- configureMCMC(R.model,
+                           print=FALSE,
+                           monitors = monitors,
+                           thin=1)
+mcmc <- buildMCMC(mcmc.spec)
+message('MCMC built')
+
+## compile model in C++
+C.model <- compileNimble(R.model)
+C.mcmc <- compileNimble(mcmc, project = R.model)
+message('NIMBLE model compiled')
+
+source('../cppp/src/calcCPPP.R', chdir = TRUE)
+options(mc.cores=6)
+
+generateCPPP(R.model,
+             C.model,
+             C.mcmc,
+             mcmc,
+             dataName = 'y',
+             paramNames = monitors, 
+             MCMCIter = 1000, 
+             NSamp = 1000,
+             NPDist = 100,
+             thin = 1)
+
+
+list(code=ss.ms.occ,
+               constants=constants,
+               data=model.data,
+               inits=inits)

### 2016-09-02 singleSpp-multiSea/opt4-5.R
cppp function working

@@ -106,7 +106,7 @@ save(ss.ms.opt5, file="saved/opt5.Rdata")
 
 
 ## *********************************************************************
-
+options(error=recover)
 ## build model
 R.model <- nimbleModel(code=ss.ms.occ,
                        constants=constants,

### 2016-09-02 singleSpp-multiSea/src/dynamicOcc.R

@@ -42,7 +42,7 @@ dbern_vec <- nimbleFunction(
 ## always treat it as 1.  It should not be used as the length needed
 rbern_vec <- nimbleFunction(
   run = function(n = integer(), p = double(), length = double()) {
-    declare(ans, double(1))
+    ans <- double(1)
     setSize(ans, length)
     for(i in 1:length) ans[i] <- rbinom(1, size = 1, prob = p)
     returnType(double(1))
@@ -86,7 +86,7 @@ dbern_matrix <- nimbleFunction(
 
 rbern_matrix <- nimbleFunction(
   run = function(n = integer(), p = double(2)) {
-    declare(ans, double(2))
+    ans = double(2)
     nrow = p[1]
     ncol = p[2]
     setSize(ans, nrow, ncol)
@@ -294,7 +294,7 @@ rDynamicOccupancy <- nimbleFunction(
     p = double(1),
     log = integer(0, default = 0)) {
     nyear <- length(p)
-    declare(ans, double(2))
+    ans <- double(2)
     setSize(ans, nrep, nyear)
     ## could populate ans here, but I'm just doing this as a placeholder
     returnType(double(2))
@@ -313,7 +313,3 @@ registerDistributions(list(
       'gamma = double(1)',
       'p = double(1)'))
   ))
-
-## *********************************************************************
-## misc functions
-## *********************************************************************

### 2016-09-09 analysis/cppp/src/calcCPPP.R
cleaned up varaible anmesing in cppp

@@ -2,7 +2,8 @@ library(parallel)
 ## draws a random number from the posterior, reassigned it as the
 ## paramter value, simulates data from the model
 pppFunc <- nimbleFunction(
-  setup = function(model, dataNames, paramNames, mcmcMV, MCMCIter, thin, burnIn){
+  setup = function(model, dataNames, paramNames,
+    mcmcMV, MCMCIter, thin, burnIn){
     paramDependencies <- model$getDependencies(paramNames)
   },
   run = function(N = integer(0)){
@@ -20,63 +21,56 @@ pppFunc <- nimbleFunction(
 
 
 
-
-generateCPPP <-  function(R.model,
-                          C.model,
-                          C.mcmc,
-                          mcmc,
-                          dataNames,
-                          paramNames, ## vector of parameters to monitor
-                          MCMCIter, ## number of samples
-                          NSamp,
-                          NPDist,
-                          burnInProportion,
-                          thin,...){
-  
-  burnIn <- ceiling(burnInProportion*(MCMCIter/thin))
-  
-  ## calculate proportion of deviances that are greater or equal to
-  ## the observed value
-  calcCPPP <- function(MCMCIter, C.model, NSamp,
-                       C.pppFunc, C.averageFunc,
-                       C.mcmc, paramNames){
-    
-    C.mcmc$run(MCMCIter)
-    observedDisc <- C.averageFunc$run()
-    if(!is.finite(observedDisc)) return(NA)
-    otherDiscs <- C.pppFunc$run(NSamp)
-    pre.pp <- mean(otherDiscs >= observedDisc)
-    if(!is.finite(pre.pp))    pre.pp <- NA
-    return(pre.pp)    
-  }
-
-  ## simulate the distribution of posterior predictive pvalues
-  simPppDist <- function(interation){
-    simulate(C.model,  includeData =  TRUE)
-    out <- calcCPPP(MCMCIter, C.model, NSamp, C.pppFunc, C.averageFunc, C.mcmc, paramNames)
-    return(out)
-  }
-  
-  
-  discAverageFunc <- nimbleFunction(
-    setup = function(model, paramNames, mcmcMV, MCMCIter, thin, burnIn){
-    },
-    run = function(){
-      mcmcSamps <-floor(MCMCIter/thin - burnIn)
-      discMean <- 0
-      for(i in 1:mcmcSamps){
-        copy(mcmcMV, model, paramNames, paramNames, row = i + burnIn)
-        discMean <- discMean + calculate(model)
-      }
[…]

### 2016-09-09 analysis/singleSpp-multiSea/opt2-3.R

@@ -43,7 +43,7 @@ ss.ms.occ <- nimbleCode({
 })
 
 input1 <- c(code=ss.ms.occ,
-           model.input)
+            model.input)
 
 ## *********************************************************************
 ## opt 2: add custom z sampler and slice on uniform(0,1) nodes
@@ -59,7 +59,7 @@ MCMCdefs.opt2 <- list('nimbleOpt2' = quote({
   zeroOneNodes <- Rmodel$getNodeNames(topOnly = TRUE)
   for(zon in zeroOneNodes) customSpec$addSampler(target = zon,
                                                  type =
-                                                   "slice",
+                                                 "slice",
                                                  print=FALSE)
   customSpec
 }))
@@ -82,52 +82,48 @@ MCMCdefs.opt2 <- list('nimbleOpt2' = quote({
 ## model assessment
 ## *********************************************************************
 ## build model
-R.model <- nimbleModel(code=ss.ms.occ,
+occ.R.model <- nimbleModel(code=ss.ms.occ,
                        constants=input1$constants,
                        data=input1$data,
                        inits=input1$inits,
                        check=FALSE)
 message('R model created')
 
-customSpec <- configureMCMC(R.model,
+customSpec <- configureMCMC(occ.R.model,
                             monitors=input1$monitors)
 customSpec$removeSamplers('phi', print=FALSE)
 customSpec$removeSamplers('gamma', print=FALSE)
 customSpec$removeSamplers('p', print=FALSE)
 customSpec$removeSamplers('psi1', print=FALSE)
 ## happens to be all top nodes
-zeroOneNodes <- R.model$getNodeNames(topOnly = TRUE)
+zeroOneNodes <- occ.R.model$getNodeNames(topOnly = TRUE)
 for(zon in zeroOneNodes) customSpec$addSampler(target = zon,
-                                                 type =
-                                                   "slice",
-                                                 print=FALSE)
+                                               type =
+                                               "slice",
+                                               print=FALSE)
 
-mcmc <- buildMCMC(customSpec)
+occ.mcmc <- buildMCMC(customSpec)
 message('MCMC built')
 
 ## compile model in C++
-C.model <- compileNimble(R.model)
-C.mcmc <- compileNimble(mcmc, project = R.model)
+occ.C.model <- compileNimble(occ.R.model)
+occ.C.mcmc <- compileNimble(occ.mcmc, project = occ.R.model)
 message('NIMBLE model compiled')
 
[…]

### 2016-09-09 ms/ms/ms.tex
spatial analysis working

--- 2016-09-01 ms/ms/ms.tex
@@
 
-\def\title{Increasing the tractability of occupancy models}
-\def\author{Lauren C.\ Ponisio$^{1,2}$, Perry de Valpine$^1$}
+\def\title{Something occupancy models}
+\def\author{Lauren C.\ Ponisio$^{1,2}$, Nicholas Michaud$^1$, Perry de
+  Valpine$^1$}
 
@@
 \begin{abstract}  
-place holder
+  \begin{enumerate}
+  \item occupancy models are everywhere, but model fitting and
+    assessment are extremely computationally intensive
+  \item because models are so computationally intensive, users often
+    forgo model assessment (determining if a model provides an
+    adequate fit to a particular dataset) because if involves
+    simulating from and refitting the model many times.
+  \item Using the NIMBLE package for R, we develop combined
+    computational approaches including user-defined and automatic
+    blocking of parameters for MCMC, filtering over latent states, and
+    customized MCMC samplers for specific parameters to improve
+    efficiency. We test these approaches using three representative
+    occupancy models of varying levels of complexity including a
+    single species model with spatial auto-correlation, a single
+    species dynamic (multi-season) model, and a multi-species
+    model. We also develop and implement methods for calculating
+    calibrated predistive posterior $p$-values to assess model fit
+    within the open source modeling software, NIMBLE.
+  \item These computation approaches lead to an improvement in MCMC
+    sampling efficiency over, particularly with models including
+    random effects.
+  \item Ours results highlight the need for more customizable
+    approaches to MCMC to fit and assess hierarchical models in order
+    to ensure occupancy models are accessible to practitioners. By
+    implementing MCMC procedures and model assessment techniques in
+    open source software, we have made progress toward this aim.
+  \item \textit{Implications:}
+  \end{enumerate}
 \end{abstract}
@@
 Estimating the proportion of sites occupied by a species is common
-challange for many subdisciplines ecology and evolution including
-metapopulation, engangered and invasion biology. Greater acceptance of
-the biases of imperfect detection has lead to the development and
-proliferation of occupancy models, which model the occurrence of a
-species at a site as a latent state layered underneath a detection
-process \citep[e.g.,][]{mackenzie-2006, royle-2007-1813}. Now only a
-little over a decade after occupancy models were introduced to
-ecology, they are being used to model the occurence of everything from
-bees \citep{mgonigle-2015-x} to tigers \citep{hines2010tigers} in an
-endless variety of complexity.
+challenge for many sub-disciplines ecology and evolution including
+meta-population, endangered species and invasion biology. Greater
+acceptance of the baises introduced by imperfect detection has lead to
+the development and proliferation of occupancy models --- models where
+the occurrence of a species at a site as a latent state layered
+underneath a detection process \citep[e.g.,][]{mackenzie-2006,
+  royle-2007-1813}. Now only a little over a decade after occupancy
+models were introduced to ecology, they are being used to model the
+occurrence of everything from bees \citep{mgonigle-2015-x} to tigers
+\citep{hines2010tigers} in an endless variety of complexity.
 
@@
 occupancy or shared variation of in occupancy across species is often
-of greatest interest to ecologsits
+of greatest interest to ecologists
 \citep[e.g.,][]{iknayan2014detecting}, the Hidden Markov Models are
-embedded in a larger hierarchical model. In such cases, practitioners
-may rely on Markov chain Monte Carlo (MCMC) to perform a Bayesian
-analysis. Such models are computationally intensive, and large models
-requiring hundreds or thousands of dimensions which require MCMC can
-be computationally impractical. To ensure occupancy models are
-accessible to practitioners, more efficient methods for estimating these
-models are necessary.
+embedded within a hierarchical model. In such cases, practitioners
+generally rely on Markov chain Monte Carlo (MCMC) to perform a
+Bayesian analysis. Standard MCMC software will including the latent
+state varaibles in MCMC sampling \citep[e.g.,][]{plummer-2003-jags,
+  winbugs, openbugs}. Such models are computationally intensive, and
+large models requiring hundreds or thousands of dimensions which
+require MCMC can be intractable. Filtering over latent states to
+calculate model likelihoods inorder to limit MCMC sampling to
+top-level parameters \citep{turek2016efficient}
 
-%% model selection
+In addition, fitting these models is such a challenge that users often
+forgo adding any additional computation to asses model fit.  A common
+idea behind evaluating whether a model provides an adequate fit to a
+particular dataset is that if data is simulated from the model, the
+simulated data should resemble the observed data. This is the basis of
+posterior predictive $P$-values, which compare the distribution of
+summary statistics calculated from simulated datasets to the observed
+statistic. Posterior predictive $P$-values alone, however, often fail
+to reject poor-fitting models \citep{bayarri-berger-00,
+  robins-etal-00, hjort-etal-06}. Methods for correcting posterior
+predictive $P$-values for better performance have been proposed
+\citep[e.g., calibrated posterior predictive $P$-values,
+][]{hjort-etal-06}, but refitting the model via MCMC iterativly.
+
+To ensure occupancy models are accessible to practitioners, more
+efficient methods for fitting and assessing these models are
+necessary.
 
 \section*{Materials \& Methods}
+
+A common idea behind evaluating whether a model provides an adequate
+fit to a particular dataset is that if data is simulated from the
+model, the simulated data should resemble the observed data. This is
+the basis of posterior predictive $P$-values, which compare the
+distribution of summary statistics calculated from simulated datasets
+to the observed statistic. Posterior predictive $P$-values alone,
+however, often fail to reject poor-fitting models
+\citep{bayarri-berger-00, robins-etal-00, hjort-etal-06}. Methods for
+correcting posterior predictive $P$-values for better performance have
+been proposed, but previously were too computationally intensive to be
+feasible \citep[e.g., calibrated posterior predictive $P$-values,
+][]{hjort-etal-06}.
+
 \label{sec:methods}

### 2016-09-09 analysis/multiSpp-singleSea/compare.R

@@ -51,5 +51,5 @@ make_MCMC_comparison_pages(ms.ss.occ.all,
 
 checkChains(ms.ss.occ.all[[1]]$samples,
             f.path = file.path(save.dir,
-            "../figures/comparisons/chains/%s.pdf")
+            "../figures/chains/%s.pdf")
 )

### 2016-09-09 analysis/spatial/compare.R

@@ -1,5 +1,7 @@
 rm(list=ls())
+library(nimble)
 setwd("~/Dropbox/nimble/occupancy/analysis/spatial")
+source('../all/plotting.R')
 save.dir <-  "../../../saved/spatial/saved"
 
 ## original model jags and nimble
@@ -11,22 +13,19 @@ load(file=file.path(save.dir, "opt1.Rdata"))
 ## custom sampler for zs, slice for other parms
 load(file=file.path(save.dir, "opt2.Rdata"))
 
-## custom sampler for zs, reflective sampler for other parms
-load(file=file.path(save.dir, "opt3.Rdata"))
-
 ## rename results
 sp.opt2[[1]] <- rename_MCMC_comparison_method('nimbleOpt2',
                                                  'slice',
                                                  comparison=sp.opt2[[1]])
-sp.opt3[[1]] <- rename_MCMC_comparison_method('nimbleOpt3',
-                                                 'reflective',
-                                                 comparison=sp.opt3[[1]])
 ## compare mcmcs
 sp.occ.all <- combine_MCMC_comparison_results(sp.orig[[1]],
                                                  sp.opt1[[1]],
                                                  sp.opt2[[1]],
-                                                 sp.opt3[[1]],
                                                  name = "sp" )
-
 make_MCMC_comparison_pages(sp.occ.all,
-                           dir=file.path(save.dir, "figures/comparisons"))
+                           dir=file.path(save.dir,
+                             "../figures/comparisons"))
+
+checkChains(sp.occ.all[[1]]$samples,
+            f.path = file.path(save.dir,
+            "../figures/chains/%s.pdf"))

### 2016-09-09 analysis/spatial/opt1-3.R

@@ -2,13 +2,14 @@ rm(list=ls())
 setwd("~/Dropbox/nimble/occupancy/analysis/spatial")
 source('src/initialize.R')
 
+set.seed(44)
 dats <- genSpatialOccData()
 model.input <- prepModData(dats$data, dats$y, dats$distance,
-                           nsite=25)
+                           nsite=50)
 
 sp.mod <- nimbleCode({
   ## priors
-  delta ~ dunif(0, 0.1)
+  delta ~ dunif(0, 5)
   sigma ~ dunif(0, 10)
   p ~ dunif(0, 1)
   alpha ~ dnorm(0, 0.001)
@@ -56,30 +57,28 @@ save(sp.opt1, file=file.path(save.dir, "opt1.Rdata"))
 ## *********************************************************************
 ## opt 2: add custom z sampler and slice on uniform(0,1) nodes
 ## *********************************************************************
+library(devtools)
+install_github("nlmichaud/nimble/packages/nimble", ref= "AFSliceSampler")
 
 MCMCdefs.opt2 <- list('nimbleOpt2' = quote({
   customSpec <- configureMCMC(Rmodel)
-  ## identify samplers to replace
-  znodes <- Rmodel$expandNodeNames('z')
-  znodes <- znodes[!Rmodel$isData(znodes)]
-  ## remove samplers
-  customSpec$removeSamplers(znodes, print=FALSE)
-  ## add custom samples
-  for(znode in znodes) customSpec$addSampler(target = znode,
-                                             type = custom_z_sampler,
-                                             print=FALSE)
   ## slice sampler for 0,1 varaibles
   customSpec$removeSamplers('p', print=FALSE)
   customSpec$addSampler(target = 'p',
                         type =
                         "slice",
                         print=FALSE)
-  ## multivariate normal sampler
-  ## customSpec$removeSamplers('rho', print=FALSE)
-  ## customSpec$addSampler(target = 'rho',
-  ##                                                type =
-  ##                                                  "ess",
-  ##                                                print=FALSE)
+  ## multivariate slice sampler
+  customSpec$removeSamplers('rho', print=FALSE)
+  customSpec$addSampler(target = 'rho',
+                                                 type =
+                                                   "AF_slice",
+                                                 print=FALSE,
+                        control = list(sliceWidths =
+                          rep(1, input1$constants$nsite),
+                          sliceFactorBurnInIters = 5000,
[…]

### 2016-09-09 analysis/spatial/original.R

@@ -4,9 +4,10 @@ source('src/initialize.R')
 library(rjags)
 load.module("msm")
 
+set.seed(444)
 dats <- genSpatialOccData()
 model.input <- prepModData(dats$data, dats$y, dats$distance,
-                           nsite=25)
+                           nsite=50)
 
 mexp <- nimbleFunction(
   run = function(A = double(2)){
@@ -17,7 +18,7 @@ mexp <- nimbleFunction(
 
 sp.mod <- nimbleCode({
   ## priors
-  delta ~ dunif(0, 10)
+  delta ~ dunif(0, 5)
   sigma ~ dunif(0, 10)
   p ~ dunif(0, 1)
   alpha ~ dnorm(0, 0.001)
@@ -71,7 +72,4 @@ sp.orig <- compareMCMCs(input1,
                         summary=FALSE,
                         check=FALSE)
 
-save(sp.orig, file="saved/orig.Rdata")
-
-checkChains(sp.orig[[1]]$samples,
-            f.path = "figures/chains/%s.pdf")
+save(sp.orig, file=file.path(save.dir, "orig.Rdata"))

### 2016-09-09 analysis/spatial/src/dataGen.R

@@ -10,12 +10,12 @@ rmvn <- function(n, mu = 0, V = matrix(1)) {
 
 
 genSpatialOccData <- function(ngrid = 25,
-                              nreps = 5,
-                              alpha = 0.1,
+                              nreps = 10,
+                              alpha = 1,
                               beta1 = 2,
                               p = 0.6,
-                              sigma = 0.5,
-                              delta = 2){
+                              sigma = 1,
+                              delta = 0.5){
 
   ## Set up a square lattice region
   simgrid <- expand.grid(1:ngrid, 1:ngrid)
@@ -24,7 +24,7 @@ genSpatialOccData <- function(ngrid = 25,
   ## Set up distance matrix
   distance <- as.matrix(dist(simgrid))
 
-  ## Generate spatial random 
+  ## Generate spatial random
   X <- rmvn(1, rep(0, n),  (sigma^2)*exp(-delta * distance))
   Xraster <- rasterFromXYZ(cbind(simgrid[, 1:2] - 0.5, X))
 

### 2016-09-09 analysis/spatial/src/initialize.R

@@ -14,6 +14,6 @@ source("src/dataGen.R")
 save.dir <-  "../../../saved/spatial/saved"
 
 ## MCMC settings
-scale <- 1e1
+scale <- 1e2
 burnin <- 1e1*scale
 niter <- (1e3)*scale

### 2016-09-09 analysis/cppp/src/calcCPPP.R
Made sure cppp calculation doesn't overwrite data.  Added easier way to read in mcmc spec.

@@ -68,6 +68,8 @@ generateCPPP <-  function(R.model,
 
   burnIn <- ceiling(burnInProportion*(MCMCIter/thin))
   
+  origData <- nimble:::values(orig.C.model, dataNames)
+  
   ## sample posterior, simulate data from sample 
   paramDependencies <- orig.C.model$getDependencies(paramNames)
   mcmcMV <- orig.mcmc$mvSamples
@@ -115,6 +117,8 @@ generateCPPP <-  function(R.model,
 
   ## calculates the number of simulated ppp that fall below the obs
   out.cppp <- mean(obs.cppp <= sim.cppp, na.rm = TRUE)  
+  
+  nimble:::values(orig.C.model, dataNames) <- origData 
   
   return(list(cppp=out.cppp,
               obs=obs.cppp,

### 2016-09-09 analysis/singleSpp-multiSea/opt2-3.R

@@ -1,4 +1,4 @@
-rm(list=ls())
+# rm(list=ls())
 setwd("~/Dropbox/nimble/occupancy/analysis/singleSpp-multiSea")
 source('src/initialize.R')
 
@@ -64,6 +64,13 @@ MCMCdefs.opt2 <- list('nimbleOpt2' = quote({
   customSpec
 }))
 
+simpSetMCMCDefs <- function(Rmodel, MCMCdefs, MCMCname) {
+  eval(MCMCdefs[[MCMCname]])
+}
+
+
+eval(MCMCdefs.opt2[['nimbleOpt2']])
+
 ## *********************************************************************
 ## run with compareMCMCs
 
@@ -87,21 +94,11 @@ occ.R.model <- nimbleModel(code=ss.ms.occ,
                        data=input1$data,
                        inits=input1$inits,
                        check=FALSE)
+
+
 message('R model created')
 
-customSpec <- configureMCMC(occ.R.model,
-                            monitors=input1$monitors)
-customSpec$removeSamplers('phi', print=FALSE)
-customSpec$removeSamplers('gamma', print=FALSE)
-customSpec$removeSamplers('p', print=FALSE)
-customSpec$removeSamplers('psi1', print=FALSE)
-## happens to be all top nodes
-zeroOneNodes <- occ.R.model$getNodeNames(topOnly = TRUE)
-for(zon in zeroOneNodes) customSpec$addSampler(target = zon,
-                                               type =
-                                               "slice",
-                                               print=FALSE)
-
+customSpec <- simpSetMCMCDefs(occ.R.model, MCMCdefs.opt2, 'nimbleOpt2')
 occ.mcmc <- buildMCMC(customSpec)
 message('MCMC built')
 
@@ -124,6 +121,4 @@ test.opt2 <- generateCPPP(occ.R.model,
                           NPDist = 5,
                           burnInProportion = 0.10,
                           thin = 1)
-
-
 

### 2016-09-10 analysis/spatial/opt1-3.R
re ran spatial model with more sites

@@ -2,7 +2,7 @@ rm(list=ls())
 setwd("~/Dropbox/nimble/occupancy/analysis/spatial")
 source('src/initialize.R')
 
-set.seed(44)
+set.seed(444)
 dats <- genSpatialOccData()
 model.input <- prepModData(dats$data, dats$y, dats$distance,
                            nsite=50)

### 2016-09-23 ms/ms/ms.tex
cross validation is not working properly, maybe...

@@
     model. We also develop and implement methods for calculating
-    calibrated predistive posterior $p$-values to assess model fit
+    calibrated predictive posterior $p$-values to assess model fit
     within the open source modeling software, NIMBLE.
@@
 meta-population, endangered species and invasion biology. Greater
-acceptance of the baises introduced by imperfect detection has lead to
+acceptance of the biases introduced by imperfect detection has lead to
 the development and proliferation of occupancy models --- models where
@@
 Bayesian analysis. Standard MCMC software will including the latent
-state varaibles in MCMC sampling \citep[e.g.,][]{plummer-2003-jags,
+state variables in MCMC sampling \citep[e.g.,][]{plummer-2003-jags,
   winbugs, openbugs}. Such models are computationally intensive, and
 large models requiring hundreds or thousands of dimensions which
-require MCMC can be intractable. Filtering over latent states to
-calculate model likelihoods inorder to limit MCMC sampling to
-top-level parameters \citep{turek2016efficient}
+require MCMC can be intractable.
 
@@
 simulated data should resemble the observed data. This is the basis of
-posterior predictive $P$-values, which compare the distribution of
+posterior predictive $p$-values, which compare the distribution of
 summary statistics calculated from simulated datasets to the observed
-statistic. Posterior predictive $P$-values alone, however, often fail
+statistic. Posterior predictive $p$-values alone, however, often fail
 to reject poor-fitting models \citep{bayarri-berger-00,
   robins-etal-00, hjort-etal-06}. Methods for correcting posterior
-predictive $P$-values for better performance have been proposed
-\citep[e.g., calibrated posterior predictive $P$-values,
-][]{hjort-etal-06}, but refitting the model via MCMC iterativly.
+predictive $p$-values for better performance have been proposed
+\citep[e.g., calibrated posterior predictive $p$-values,
+][]{hjort-etal-06}, but refitting the model via MCMC iterativly. Given
+that with occupancy models fitting the model just once can be a time
+consuming task, efficient methods for MCMC are necessary to ensure
+methods for assessment are feasible for these models.
 
-To ensure occupancy models are accessible to practitioners, more
-efficient methods for fitting and assessing these models are
-necessary.
+Beyond assessing the fit of a model, choosing between models is one of
+the most widely used applications of statistics by
+practitioners. Though many theoretically sound methods for Bayesian
+model selection such as cross-validation have been developed
+\citep{hooten2014guide}, they, like model assessment, are
+computationally intensive --- particularly for hierarchical models
+like occupancy models. A typical need for model selection arises when
+a practitioner is choosing whether to include a specific layer of
+hierarchy (i.e., random effect). This is often the case with so called
+``multi-species'' occupancy models, where the occupancy of many
+species is estimated simultaneously in a model with a random effect of
+species \citep[reviewed in, ][]{iknayan2014detecting}. Ecologists are
+often interested in whether there is some variability in the response
+of species to an explanatory variable such that a random effect of
+species accounts for that variability, or whether a fixed effect of
+that variable fits adequately
+\citep{pacifici2014guidelines}. Currently, the Deviance Information
+Criteria (DIC), originally derived to mimic AIC for Bayesian,
+non-hierarchical models, is now commonly used by scientists to
+evaluate hierarchical models. Though the limitations of DIC for
+hierarchical model selection are widely recognized by statisticians
+\citep{celeux2006deviance, hooten2014guide}, because it is built into
+open-source software such as WinBUGS \citep{winbugs}, it is
+uncritically used by practitioners. Readily available and
+theoretically sound alternative methods are thus critically needed.
+
+% Filtering over latent states to calculate model likelihoods inorder to
+% limit MCMC sampling to top-level parameters dyanmic blocking or
+% parameters \citep{turek2016efficient}
 
 \section*{Materials \& Methods}
+\label{sec:methods}
+\subsection*{Computational approaches}
 
-A common idea behind evaluating whether a model provides an adequate
-fit to a particular dataset is that if data is simulated from the
-model, the simulated data should resemble the observed data. This is
-the basis of posterior predictive $P$-values, which compare the
-distribution of summary statistics calculated from simulated datasets
-to the observed statistic. Posterior predictive $P$-values alone,
-however, often fail to reject poor-fitting models
-\citep{bayarri-berger-00, robins-etal-00, hjort-etal-06}. Methods for
-correcting posterior predictive $P$-values for better performance have
-been proposed, but previously were too computationally intensive to be
-feasible \citep[e.g., calibrated posterior predictive $P$-values,
-][]{hjort-etal-06}.
+\subsubsection*{Single species, single season occupancy model with
+  spatial auto-correlation}
+The first model is a single species, single season occupancy model
+accounting for spatial auto-correlation. We let $z_{i}$ denote the
+true occupancy of a species at site $i$.  We then let $x_{i,j}$
+indicate whether we detected ($x_{i,j}=1$) or did not detect
+($x_{i,j}=0$) that species in the $j^{\mathrm{th}}$ visit to site $i$.
+We assumed that occupancy at the $i^{\mathrm{th}}$ site is a Bernoulli
+random variable $z_{i} \sim \mathrm{Bern}(\psi_{i})$ with probability
+$\psi_{i}$.  We included the effect of an arbitrary covariate (e.g.,
+elevation) on site occupancy. To model the spatial auto-correlation in
+occupancy between sites, we assume the co-variance between sites $Y_i$
+and $Y_j$ is a function of distance between $p_i$ and $p_j$. We
+computed the probability of occupancy at site $i$
+%
+\begin{equation}
+  \begin{split}
+    \label{eq:spatial}
+    \mathrm{logit}(\psi_{i}) &=
+    \alpha + \beta*elevation_i + \rho_i\hspace{0.2em}\\
+    \rho_i &\sim MVN(0, {Cov}(Y_i,Y_j)) \hspace{0.2em}\\
+    {Cov} (Y_i,Y_j) &= \sigma^2exp^{(-\lambda\|p_i-p_j\|)} \hspace{0.2em}.\\
+  \end{split}
+\end{equation}
+%
 
-\label{sec:methods}
+Where $\lambda$ is the exponential decay constant and $\sigma^2$ is
+SOMETHING....
+
+To improve efficiency of this model...
+
+\subsubsection*{Single species, multi season (dynamic) occupancy
+  model}
+\label{sec:ssms}
+
+The second model is a relatively simple single species occupancy model
+over multiple seasons \citep{Royle2007}. We let $z_{i,t}$ denote the
+true occupancy of a species in year $t$ at site $i$.  We then let
+$x_{i,t,j}$ indicate whether we detected ($x_{i,t,j}=1$) or did not
+detect ($x_{i,t,j}=0$) that species in the $j^{\mathrm{th}}$ visit to
+site $i$ in year $t$.  We assumed that occupancy at the
+$i^{\mathrm{th}}$ site in the $t^{\mathrm{th}}$ year is a Bernoulli
+random variable $z_{i,t} \sim \mathrm{Bern}(\psi_{i,t})$ with
+probability $\psi_{i,t}$.
+
+Letting $\phi_{i,t}$ denote the probability the species persists at
+site $i$ from years $t$ to $t+1$ (provided it was present at site $j$
+in year $t$, $z_{i,t}=1$) and $\gamma_{i,t}$ denote the
+probability that site $i$ is colonized in year $t+1$
+(provided it was not present at site $i$ in year $t$, $z_{i,t}=0$),
+we then computed the probability of occupancy  at site
+$i$ in subsequent years as
+%
+\begin{equation}
+  \label{eq:occupancy2}
+  \psi_{i,t+1} =
+  \phi_{i,t} * z_{i,t} + \gamma_{i,t} * (1-z_{i,t})\hspace{0.2em}.
+\end{equation}
+%
+
+First, to improve efficiency, filter over latent states
+to calculate model likelihoods in order to limit MCMC sampling to
+top-level parameters. We then use two computational approaches to
+improve the efficiency of this model 1) dynamic blocking of the
+parameters \citep{turek2016efficient}, and 2) a custom MCMC
+specification where a slice sampler is used for all parameters.
+
+\subsubsection*{Multi species, single season occupancy model}
+\label{sec:msss}
+
+The last model is from \citep{zipkin2010multi}. It is a multi-species,
+single season occupancy model examining the effect of wildlife
+management and habitat characteristics on bird communities
+\citep{zipkin2010multi}. The species-specific coefficients for the
+effect of basal tree area, understory foliage and deer management
+where bound together by a common distribution with an estimated
+variance. The model is similar to Eq.~\ref{eq:spatial}, except each for
+species $k$, we let $z_{i,k}$ denote its true occupancy state at site
+$i$.
+
+To improve the efficiency of this model, we first filtered over latent
+states to calculate model likelihoods in order to limit MCMC sampling
+to top-level parameters. We also vectorized all calculations that
+would have require for loops in BUGS or JAGS. We then tried two
+approaches to speed sampling of the top-level parameters 1) dynamic
+blocking of the parameters \citep{turek2016efficient}, and 2) a custom
+blocking scheme were the parameters of each species are blocked
+together.
+
+\subsection*{Model assessment}
+\label{sec:assess}
+
+We implemented a procedure to calculate calibrated posterior
+predictive $p$-values \citep{hjort-etal-06}. After the parameters have
+been fit to the model, a sample of the posterior is used to simulate
+data from the model. A discrepancy measure, which we chose to be the
+model likelihood, is then calculated, and the posterior $p$-value is
+the number of simulated $p$-values that fall below the observed. To
+``calibrate'' the distribution of posterior $p$-values, the MCMC is
+rerun on the simulated data to refit the model.
+
+\subsection*{Model selection}
+\label{sec:select}
+
+Cross-validation is one of the most fudamental procedures in model
+selection, but, because it requires iternativly re-fitting the model,
+is computationally intensive \citep{hooten2014guide}. In cross
+validation, we exclude a subset of the data ($y_k$) from model
+fitting, then use the fitted model to predict $y_k$. The preidtion
+error is summarized by coparing the simulated $y_k$ to the true $y_k$.
+
+In the multi species occupancy models (Section \ref{sec:msss}),
+practitioners are often interested in determining whether a model
+including a species random effect for explanatory variables is a
+better fit than a model without the random effect. We implemented a
+cross-validation procudure for this model where the detection data for
+species is left out, the model refitted, and the fitted model used to
+predict the occurence of that species. The predictive error of the
+model included a random effect of species is then compared to a model
+where no species random effects were included.
 

### 2016-09-23 analysis/crossValidation/crossValidationFunction.R

@@ -2,36 +2,6 @@ rm(list=ls())
 library(nimble)
 library(parallel)
 
-dyesCode <- nimbleCode({
-  for (i in 1:BATCHES) {
-    for (j in 1:SAMPLES) {
-      y[i,j] ~ dnorm(mu[i], sd = sigma.within);
-    }
-    mu[i] ~ dnorm(theta, sd = sigma.between);
-  }
-  
-  theta ~ dnorm(0.0, 1.0E-10);
-  sigma.within ~ dunif(0, 100)
-  sigma.between ~ dunif(0, 100)
-})
-
-dyesModel <- nimbleModel(dyesCode,
-                         constants = list(BATCHES = 6, SAMPLES = 5))
-
-data <- matrix(c(1545, 1540, 1595, 1445, 1595, 1520, 1440, 1555, 1550,
-                 1440, 1630, 1455, 1440, 1490, 1605, 1595, 1515, 1450,
-                 1520, 1560, 1510, 1465, 1635, 1480, 1580, 1495, 1560,
-                 1545, 1625, 1445), nrow = 6)
-
-dyesModel$setData(list(y = data))
-
-## crossValCalculate <- function(row, MCMCOut, dataDimensions, saveData){
-##   simDataArray <- array(MCMCOut[row,], dim = c(dataDimensions))
-##   evalCode0 <- paste0("simData <- simDataArray[",rep(",", leaveOutIndex - 1), i, paste0(rep(",", length(dataDimensions) - leaveOutIndex)),"]")
-##   eval(parse(text = evalCode0))
-##   discrep <- simData - saveData
-##   return(discrep)
-## }
 
 crossValCalculate <- function(row, MCMCOut, dataDimensions, saveData){
   simDataArray <- array(MCMCOut[row,], dim = c(dataDimensions))
@@ -39,22 +9,20 @@ crossValCalculate <- function(row, MCMCOut, dataDimensions, saveData){
   return(discrep)
 }
 
-
 crossValidateOne <- function(model,
                              dataNames,
                              MCMCIter,
                              burnIn, thin,
-                             leaveOutIndex
+                             leaveOutIndex,
                              MCMCdefs=NULL){
   simpSetMCMCDefs <- function(Rmodel, MCMCdefs, MCMCname) {
     eval(MCMCdefs[[MCMCname]])
   }
-  if(!is.null(MCMCdefs){
[…]

### 2016-09-23 analysis/multiSpp-singleSea/original.R

@@ -137,6 +137,14 @@ model.input$constants[["max.num.reps"]] <- NULL
 input1 <- c(code=ms.ss.occ, model.input)
 
 
+occ.R.model <- nimbleModel(code=ms.ss.occ,
+                           constants=input1$constants,
+                           data=input1$data,
+                           inits=input1$inits,
+                           check=FALSE)
+
+
+
 ## *********************************************************************
 ## original model with nimble
 

### 2016-09-23 analysis/singleSpp-multiSea/original.R

@@ -42,14 +42,14 @@ ss.ms.occ <- nimbleCode({
 
   ## Derived parameters: Sample and population occupancy, growth rate
   ## and turnover
-  psi[1] <- psi1
-  n.occ[1]<-sum(z[1:nsite,1])
-  for (k in 2:nyear){
-    psi[k] <- psi[k-1]*phi[k-1] + (1-psi[k-1])*gamma[k-1]
-    n.occ[k] <- sum(z[1:nsite,k])
-    growthr[k-1] <- psi[k]/psi[k-1]
-    turnover[k-1] <- (1 - psi[k-1]) * gamma[k-1]/psi[k]
-  }
+  ## psi[1] <- psi1
+  ## n.occ[1]<-sum(z[1:nsite,1])
+  ## for (k in 2:nyear){
+  ##   psi[k] <- psi[k-1]*phi[k-1] + (1-psi[k-1])*gamma[k-1]
+  ##   n.occ[k] <- sum(z[1:nsite,k])
+  ##   growthr[k-1] <- psi[k]/psi[k-1]
+  ##   turnover[k-1] <- (1 - psi[k-1]) * gamma[k-1]/psi[k]
+  ## }
 })
 
 ## *********************************************************************

### 2016-10-11 ms/ms/ms.tex
various bug fixes

@@
 parameters \citep{turek2016efficient}, and 2) a custom MCMC
-specification where a slice sampler is used for all parameters.
+specification where slice samplers \cite{neal-03} are used for all parameters.
+
+Slice samplers are a class of methods that sample from a target distribution  by using that fact that samples from any distribution can be obtained by sampling uniformly from the area under that distribution's probability density function curve.
+The horizontal coordinates of these uniform samples will provide samples from the distribution of interest.
+Slice samplers have been shown to perform well in situations where choosing a tuning parameter for a Metropolis algorithm is difficult.
+
+%probably TMI below
+When used to sample from the posterior distribution of a univariate parameter, a slice sampler proceeds at each iteration by first choosing a vertical coordinate sampled uniformly between 0 and the height of the density curve at the parameter value from the previous iteration.
+Then, a horizontal coordinate is chosen uniformly from the set of all possible parameter values whose density is at least as great as the chosen vertical coordinate.
 
@@
 blocking of the parameters \citep{turek2016efficient}, and 2) a custom
-blocking scheme were the parameters of each species are blocked
-together.
+blocking scheme where the parameters of each species are blocked
+together with adaptive random walk MCMC.
+
+Random walk Metropolis algorithms \citep{metropolis1953equation} sample from the posterior distribution of a parameter of interest by first proposing a new parameter value from a normal proposal distribution centered at the current value, and then deciding whether to accept or reject the proposed value via the Metropolis ratio.
+Although common, standard univariate or multivariate normal proposal distributions can prove to be inefficient at sampling for models in which parameters are highly correlated.
+Adaptive random walk Metropolis sampling \citep{haario98anadaptive} allows the correlation structure of the posterior distribution of blocks of parameters to be used to produce better proposals.
+For an occupancy model where parameters are highly correlated within species, this can provide much more efficient sampling than a non-adaptive Metropolis algorithm.
 
@@
 validation, we exclude a subset of the data ($y_k$) from model
-fitting, then use the fitted model to predict $y_k$. The preidtion
-error is summarized by coparing the simulated $y_k$ to the true $y_k$.
+fitting, then use the fitted model to predict $y_k$. The prediction
+error is summarized by comparing the simulated $y_k$ to the true $y_k$.
 
@@
 \label{sec:results}
+\subsubsection*{Single species, single season occupancy model with
+  spatial auto-correlation}
+
+\subsubsection*{Multi species, single season occupancy model}
+\label{sec:msss}
+
+\subsubsection*{Single species, multi season (dynamic) occupancy
+  model}
 

### 2016-10-11 analysis/cppp/src/calcCPPP.R

@@ -1,10 +1,11 @@
 library(parallel)
+library(coda)
 ## draws a random number from the posterior, reassigned it as the
 ## paramter value, simulates data from the model
 
 virtualDiscFunction <- nimbleFunctionVirtual(run = function()
-  returnType(double(0))
-)
+                                             returnType(double(0))
+                                             )
 
 likeDiscFuncGenerator <- nimbleFunction(
   setup = function(model, ...){},
@@ -14,7 +15,7 @@ likeDiscFuncGenerator <- nimbleFunction(
     return(output)
   },
   contains = virtualDiscFunction
-)
+  )
 
 maxDiscFuncGenerator <- nimbleFunction(
   setup = function(model, ...){
@@ -27,7 +28,7 @@ maxDiscFuncGenerator <- nimbleFunction(
     return(output)
   },
   contains = virtualDiscFunction
-)
+  )
 
 
 pppFunc <- nimbleFunction(
@@ -83,12 +84,17 @@ calcCPPP <- function(MCMCIter,
                      C.pppFunc,
                      cppp.C.mcmc,
                      firstRun){
-  if(firstRun  == 0)
+  if(firstRun  == 0){
     cppp.C.mcmc$run(MCMCIter)
+    samples <- mcmc(as.matrix(cppp.C.mcmc$mvSamples))
+  } else{
+    samples <- NA
+  }
   
   pre.pp <- C.pppFunc$run(NSamp)
   if(!is.finite(pre.pp))    pre.pp <- NA
-  return(pre.pp)    
+  return(list(pre.pp = pre.pp,
+              samples = samples))    
 }
 
 
@@ -104,7 +110,8 @@ generateCPPP <-  function(R.model,
                           burnInProportion, ## proportion of mcmc to drop
                           thin,## thinning used in original mcmc run
                           discFuncGenerator, 
-                          averageParams, 
+                          averageParams,
+                          returnChains = TRUE,
                           ...){
   burnIn <- ceiling(burnInProportion*(MCMCIter/thin))
   origData <- nimble:::values(orig.C.model, dataNames)
@@ -132,7 +139,6 @@ generateCPPP <-  function(R.model,
                        C.pppFunc,
                        orig.C.mcmc,
                        firstRun = 1)
-  print(obs.cppp)
 
[…]

### 2016-10-11 analysis/crossValidation/crossValidationFunction.R

@@ -1,7 +1,5 @@
-rm(list=ls())
 library(nimble)
 library(parallel)
-
 
 crossValCalculate <- function(row, MCMCOut, dataDimensions, saveData){
   simDataArray <- array(MCMCOut[row,], dim = c(dataDimensions))
@@ -32,20 +30,25 @@ crossValidateOne <- function(model,
   compileNimble(newModel)
   numBlocks <- newModel$getVarInfo(dataNames)[['maxs']][leaveOutIndex]
   dataDimensions <- newModel$getVarInfo(dataNames)[['maxs']]
-  saveData <- array(nimble:::values(newModel, dataNames), dim = c(dataDimensions))
+  saveData <- array(nimble:::values(newModel, dataNames),
+                    dim = c(dataDimensions))
 
-  calcCrossVal <- function(i){
+  calcCrossVal <- function(i,
+                           saveData,
+                           newModel,
+                           leaveOutIndex,
+                           dataDimensions,
+                           dataNames){
     tempData <- saveData
-    print(tempData)
+    compileNimble(newModel)
     newModel$resetData()
-    evalCode1 <- paste0("tempData[",rep(",", leaveOutIndex - 1), i,
-                        paste0(rep(",", length(dataDimensions) -
-                                   leaveOutIndex)),"] <- NA")
+    evalCode1 <- paste0("tempData[",paste(rep(",", leaveOutIndex - 1),
+    collapse=""), i, paste(rep(",", length(dataDimensions) -
+                                   leaveOutIndex), collapse=""),"] <- NA")
     eval(parse(text = evalCode1))
     evalCode2 <- paste0("modelDataList <- list(", dataNames, "= tempData)")
     eval(parse(text=evalCode2))
     newModel$setData(modelDataList)
-    print(modelDataList)
     modelMCMCConf <- configureMCMC(newModel,
                                    monitors = dataNames, thin = thin)
     modelMCMC <- buildMCMC(modelMCMCConf)
@@ -59,56 +62,15 @@ crossValidateOne <- function(model,
                                      crossValCalculate, MCMCout,
                                      dataDimensions, saveData))
     crossValAverage <- log(mean(crossValValue))
[…]

### 2016-10-11 analysis/multiSpp-singleSea/opt1-4.R

@@ -255,7 +255,7 @@ occ.R.model <- nimbleModel(code=ms.ss.occ,
 occ.mcmc <- buildMCMC(occ.R.model)
 occ.C.model <- compileNimble(occ.R.model)
 occ.C.mcmc <- compileNimble(occ.mcmc, project = occ.R.model)
-occ.C.mcmc$run(10^3)
+occ.C.mcmc$run(niter)
 
 source('../cppp/src/calcCPPP.R', chdir = TRUE)
 options(mc.cores=6)
@@ -266,12 +266,114 @@ test.opt2 <- generateCPPP(occ.R.model,
                           occ.mcmc,
                           dataName = 'X',
                           paramNames = input1$monitors, 
-                          MCMCIter = 10^5, 
-                          NSamp = 10^4,
-                          NPDist = 10^2,
+                          MCMCIter = niter, 
+                          NSamp = 10^3,
+                          NPDist = 10^3,
                           burnInProportion = 0.10,
                           thin = 1,
                           averageParams = TRUE,
-                          discFuncGenerator=likeDiscFuncGenerator)
+                          discFuncGenerator=likeDiscFuncGenerator,
+                          returnChains=TRUE)
 
 save(test.opt2, file=file.path(save.dir, "ms_ss_noz_CPPP.Rdata"))
+
+
+## *********************************************************************
+## ## cross validation
+## *********************************************************************
+## if not already run above
+occ.R.model <- nimbleModel(code=ms.ss.occ,
+                           constants=input1$constants,
+                           data=input1$data,
+                           inits=input1$inits,
+                           check=FALSE)
+
+options(mc.cores=6)
+
+source('../crossValidation/crossValidationFunction.R')
+
+output <- crossValidateOne(model=occ.R.model,
+                           dataNames= "X",
+                           MCMCIter= niter,
+                           burnIn=niter*0.1,
+                           thin=1,
+                           leaveOutIndex=3,
+                           MCMCdefs=NULL)
+
+## *********************************************************************
+
+ms.ss.occ.simp <- nimbleCode({
+
+  ## random effects  
+  a1 ~ dnorm(0, 0.001)  
+  a2 ~ dnorm(0, 0.001)
+  a3 ~ dnorm(0, 0.001)
+  a4 ~ dnorm(0, 0.001)
[…]

### 2016-10-11 analysis/multiSpp-singleSea/original.R

@@ -136,15 +136,6 @@ model.input$constants[["max.num.reps"]] <- NULL
 
 input1 <- c(code=ms.ss.occ, model.input)
 
-
-occ.R.model <- nimbleModel(code=ms.ss.occ,
-                           constants=input1$constants,
-                           data=input1$data,
-                           inits=input1$inits,
-                           check=FALSE)
-
-
-
 ## *********************************************************************
 ## original model with nimble
 

### 2016-10-11 analysis/multiSpp-singleSea/src/reformatData.R

@@ -188,7 +188,7 @@ prepMutiSpData <- function(survey.data,
   ## initial values
   omega.draw <- runif(1, num.species/(num.species + n.zeroes), 1)
 
-  ## inital conditions with 1s as NAs and Nas as 1s
+  ## inital conditions with 1s as NAs and NAs as 1s
   zinits <- zs
   zinits[zinits == 1] <- 2
   zinits[is.na(zinits)] <- 1

### 2016-10-11 analysis/spatial/opt1-3.R

@@ -9,7 +9,7 @@ model.input <- prepModData(dats$data, dats$y, dats$distance,
 
 sp.mod <- nimbleCode({
   ## priors
-  delta ~ dunif(0.5, 10)
+  delta ~ dunif(0.01, 10)
   sigma ~ dunif(0, 10)
   p ~ dunif(0, 1)
   alpha ~ dnorm(0, 0.001)

### 2016-10-11 analysis/spatial/original.R

@@ -18,12 +18,16 @@ mexp <- nimbleFunction(
 
 sp.mod <- nimbleCode({
   ## priors
-  delta ~ dunif(0.5, 1.5)
+  delta ~ dunif(0.01, 10)
   sigma ~ dunif(0, 10)
   p ~ dunif(0, 1)
   alpha ~ dnorm(0, 0.001)
   b1 ~ dnorm(0, 0.001)
 
+  rho[1:nsite] ~ dmnorm(zeros[1:nsite],
+                        D.tau[1:nsite, 1:nsite])
+
+  
   ## Likelihood
   ## Ecological model for true occurrence
   for (i in 1:nsite) {
@@ -38,22 +42,19 @@ sp.mod <- nimbleCode({
     }
   }
 
-  rho[1:nsite] ~ dmnorm(zeros[1:nsite],
-                        D.tau[1:nsite, 1:nsite])
-
   ## create covariance matrix based on distances (must be 1/cov for
   ## JAGS)
 
-  ## mexp is jags's version fo matrix exponentiation, veyr sensitive 
+  ## mexp is jags's version fo matrix exponentiation, very sensitive 
   ## temp.cov[1:nsite, 1:nsite] <- -delta*D[1:nsite, 1:nsite]
-  ## D.cov[1:nsite, 1:nsite]  <- (sigma^2)*mexp(-delta*D[1:nsite, 1:nsite])
+  D.cov[1:nsite, 1:nsite]  <- (sigma^2)*mexp(-delta*D[1:nsite, 1:nsite])
 
-  for(i in 1:nsite){
-    for(j in 1:nsite){
-      temp.cov[i, j] <- -delta*D[i, j]
-      D.cov[i, j]  <- (sigma^2)* exp(temp.cov[i, j])
-    }
-  }
+  ## for(i in 1:nsite){
+  ##   for(j in 1:nsite){
+  ##     temp.cov[i, j] <- -delta*D[i, j]
+  ##     D.cov[i, j]  <- (sigma^2)* exp(temp.cov[i, j])
+  ##   }
+  ## }
   
   D.tau[1:nsite, 1:nsite] <- inverse(D.cov[1:nsite, 1:nsite])
   

### 2016-10-11 analysis/spatial/spatial_ex.R

@@ -24,14 +24,14 @@ rmvn <- function(n, mu = 0, V = matrix(1)) {
 
 ## simulate detection with covariate and spatial random effect data
 ## set
-genSpatialOccData <- function(ngrid = 25, ## number of spatial grid cells
+genSpatialOccData <- function(ngrid = 50, ## number of spatial grid cells
                               nreps = 10, ## number of repeated
                               ## samples at sites
                               alpha = 1, ## intercept 
                               beta1 = 2, ## covariate coefficient
                               p = 0.6, ## prob detection
-                              sigma = 1, ## 
-                              delta = 0.5){
+                              sigma = 0.5, 
+                              delta = 0.5){ ## decay parameter
 
   ## Set up a square lattice region
   simgrid <- expand.grid(1:ngrid, 1:ngrid)
@@ -125,7 +125,7 @@ model.input <- prepModData(dats$data, dats$y, dats$distance,
 ## *********************************************************************
 sp.mod <- nimbleCode({
   ## priors
-  delta ~ dunif(0, 10)
+  delta ~ dunif(0.01, 10)
   sigma ~ dunif(0, 10)
   p ~ dunif(0, 1)
   alpha ~ dnorm(0, 0.001)
@@ -163,7 +163,7 @@ input1 <- c(code=sp.mod,
 ## *********************************************************************
 
 sp.opt1 <- compareMCMCs(input1,
-                        MCMCs=c("nimble", "autoBlock"),
+                        MCMCs=c("nimble"),
                         niter=niter,
                         burnin = burnin,
                         summary=FALSE,
@@ -188,7 +188,7 @@ mexp <- nimbleFunction(
 
 sp.mod <- nimbleCode({
   ## priors
-  delta ~ dunif(0, 5)
+  delta ~ dunif(0.01, 10)
   sigma ~ dunif(0, 10)
   p ~ dunif(0, 1)
   alpha ~ dnorm(0, 0.001)
@@ -217,10 +217,9 @@ sp.mod <- nimbleCode({
   ## mexp is jags's version fo matrix exponentiation, for some reason
   ## the below will not work
   
-  ## temp.cov[1:nsite, 1:nsite] <- -delta*D[1:nsite, 1:nsite]
   ## D.cov[1:nsite, 1:nsite]  <- (sigma^2)*mexp(-delta*D[1:nsite, 1:nsite])
 
[…]

### 2016-10-11 analysis/spatial/src/dataGen.R

@@ -9,13 +9,13 @@ rmvn <- function(n, mu = 0, V = matrix(1)) {
 }
 
 
-genSpatialOccData <- function(ngrid = 25,
+genSpatialOccData <- function(ngrid = 50,
                               nreps = 10,
                               alpha = 1,
                               beta1 = 2,
                               p = 0.6,
-                              sigma = 1,
-                              delta = 1){
+                              sigma = 0.5,
+                              delta = 0.5){
 
   ## Set up a square lattice region
   simgrid <- expand.grid(1:ngrid, 1:ngrid)

### 2016-10-11 ms/ms/ms.tex
various bug fixes

@@
 \def\author{Lauren C.\ Ponisio$^{1,2}$, Nicholas Michaud$^1$, Perry de
-  Valpine$^1$}
+  Valpine$^1$ (Daniel and Chris?)}
 
@@
     assessment are extremely computationally intensive
-  \item because models are so computationally intensive, users often
+  \item Because models are so computationally intensive, users often
     forgo model assessment (determining if a model provides an
-    adequate fit to a particular dataset) because if involves
-    simulating from and refitting the model many times.
-  \item Using the NIMBLE package for R, we develop combined
-    computational approaches including user-defined and automatic
-    blocking of parameters for MCMC, filtering over latent states, and
-    customized MCMC samplers for specific parameters to improve
-    efficiency. We test these approaches using three representative
-    occupancy models of varying levels of complexity including a
-    single species model with spatial auto-correlation, a single
-    species dynamic (multi-season) model, and a multi-species
+    adequate fit to a particular dataset) and model selection
+    (choosing the best model out of a set of models) --- methods that
+    generally involve simulating from and refitting the model
+    iterativly.
+  \item Using the open source modeling software NIMBLE, we develop
+    combined computational approaches including user-defined and
+    automatic blocking of parameters for MCMC, filtering over latent
+    states, and customized MCMC samplers for specific parameters to
+    improve efficiency. We test these approaches using three
+    representative occupancy models of varying levels of complexity
+    including a single species model with spatial auto-correlation, a
+    single species dynamic (multi-season) model, and a multi-species
     model. We also develop and implement methods for calculating
-    calibrated predictive posterior $p$-values to assess model fit
-    within the open source modeling software, NIMBLE.
+    calibrated predictive posterior $p$-values to assess model fit and
+    cross validation for model selection within NIMBLE.
   \item These computation approaches lead to an improvement in MCMC
     sampling efficiency over, particularly with models including
-    random effects.
-  \item Ours results highlight the need for more customizable
-    approaches to MCMC to fit and assess hierarchical models in order
-    to ensure occupancy models are accessible to practitioners. By
-    implementing MCMC procedures and model assessment techniques in
-    open source software, we have made progress toward this aim.
-  \item \textit{Implications:}
+    random effects. (more results once they are available)
+  \item \textit{Implications:} Ours results highlight the need for
+    more customizable approaches to MCMC to fit and assess
+    hierarchical models in order to ensure occupancy and other
+    hierarchical models are accessible to practitioners. By
+    implementing MCMC procedures and model assessment and selection
+    techniques in open source software, we have made progress toward
+    this aim.
   \end{enumerate}
@@
 acceptance of the biases introduced by imperfect detection has lead to
-the development and proliferation of occupancy models --- models where
-the occurrence of a species at a site as a latent state layered
+the development and proliferation of occupancy models where the
+occurrence of a species at a site is modeled as a latent state layered
 underneath a detection process \citep[e.g.,][]{mackenzie-2006,
@@
 Occupancy models are part of a larger class of models known as Hidden
-Markov Models. For discrete Hidden Markov Models like occupancy models
+Markov Models. In discrete Hidden Markov Models like occupancy models
 where a species is either present or absent from a site, likelihood
@@
 states. Because estimating the effect of explanatory variables on site
-occupancy or shared variation of in occupancy across species is often
-of greatest interest to ecologists
-\citep[e.g.,][]{iknayan2014detecting}, the Hidden Markov Models are
-embedded within a hierarchical model. In such cases, practitioners
-generally rely on Markov chain Monte Carlo (MCMC) to perform a
-Bayesian analysis. Standard MCMC software will including the latent
-state variables in MCMC sampling \citep[e.g.,][]{plummer-2003-jags,
-  winbugs, openbugs}. Such models are computationally intensive, and
-large models requiring hundreds or thousands of dimensions which
-require MCMC can be intractable.
+occupancy or shared variation in occupancy across species is often of
+greatest interest to ecologists \citep[e.g.,][]{iknayan2014detecting},
+the Hidden Markov Model is often embedded within a hierarchical
+model. In such cases, practitioners generally rely on Markov chain
+Monte Carlo (MCMC) to perform a Bayesian analysis. Standard MCMC
+software will including the latent state variables in MCMC sampling
+\citep[e.g.,][]{plummer-2003-jags, winbugs, openbugs}. Such models are
+computationally intensive, and large models requiring hundreds or
+thousands of dimensions which require MCMC can be intractable.
 
@@
 idea behind evaluating whether a model provides an adequate fit to a
-particular dataset is that if data is simulated from the model, the
-simulated data should resemble the observed data. This is the basis of
-posterior predictive $p$-values, which compare the distribution of
-summary statistics calculated from simulated datasets to the observed
+dataset is that if data is simulated from the model, the simulated
+data should resemble the observed data. This is the basis of posterior
+predictive $p$-values, which compare the distribution of summary
+statistics calculated from simulated datasets to the observed
 statistic. Posterior predictive $p$-values alone, however, often fail
@@
   robins-etal-00, hjort-etal-06}. Methods for correcting posterior
-predictive $p$-values for better performance have been proposed
-\citep[e.g., calibrated posterior predictive $p$-values,
-][]{hjort-etal-06}, but refitting the model via MCMC iterativly. Given
-that with occupancy models fitting the model just once can be a time
-consuming task, efficient methods for MCMC are necessary to ensure
-methods for assessment are feasible for these models.
+predictive $p$-values for better performance by refitting the model
+via MCMC iterativly have been proposed \citep[e.g., calibrated
+posterior predictive $p$-values, ][]{hjort-etal-06}, but no methods
+are available in open source software. In addition, given fitting an
+occupancy model just once can be a time consuming task, efficient
+methods for MCMC are necessary to ensure methods for assessment are
+feasible for these models.
 
@@
 the most widely used applications of statistics by
-practitioners. Though many theoretically sound methods for Bayesian
-model selection such as cross-validation have been developed
-\citep{hooten2014guide}, they, like model assessment, are
-computationally intensive --- particularly for hierarchical models
-like occupancy models. A typical need for model selection arises when
-a practitioner is choosing whether to include a specific layer of
-hierarchy (i.e., random effect). This is often the case with so called
-``multi-species'' occupancy models, where the occupancy of many
-species is estimated simultaneously in a model with a random effect of
-species \citep[reviewed in, ][]{iknayan2014detecting}. Ecologists are
-often interested in whether there is some variability in the response
-of species to an explanatory variable such that a random effect of
-species accounts for that variability, or whether a fixed effect of
-that variable fits adequately
-\citep{pacifici2014guidelines}. Currently, the Deviance Information
-Criteria (DIC), originally derived to mimic AIC for Bayesian,
-non-hierarchical models, is now commonly used by scientists to
-evaluate hierarchical models. Though the limitations of DIC for
+practitioners. Though many methods for Bayesian model selection such
+as cross-validation have been developed \citep{hooten2014guide}, they,
+like model assessment, are computationally intensive --- particularly
+for hierarchical models like occupancy models. A typical need for
+model selection arises when a practitioner is choosing whether to
+include a specific layer of hierarchy (i.e., random effect). This is
+often the case with so called ``multi-species'' occupancy models,
+where the occupancy of many species is estimated simultaneously in a
+model with a random effect of species \citep[reviewed in,
+][]{iknayan2014detecting}. Ecologists are often interested in whether
+there is some variability in the response of species to an explanatory
+variable such that a random effect of species accounts for that
+variability \citep{pacifici2014guidelines}. Currently, the Deviance
+Information Criteria (DIC), originally derived to mimic AIC for
+Bayesian, non-hierarchical models, is now commonly used by scientists
+to evaluate hierarchical models. Though the limitations of DIC for
 hierarchical model selection are widely recognized by statisticians
@@
 
-% Filtering over latent states to calculate model likelihoods inorder to
-% limit MCMC sampling to top-level parameters dyanmic blocking or
-% parameters \citep{turek2016efficient}
+Luckily, methods to improve MCMC efficiency of Hidden Markov Models,
+such filtering over latent states to calculate model likelihoods
+in order to limit MCMC sampling to top-level parameters dynamic
+blocking or parameters \citep{turek2016efficient}, have been
+developed. A synergistic strategy is to assign specific MCMC samplers
+to different parameters depending on the nature of those nodes (i.e.,
+discrete versus continuous). Though these methods are available in
+isolation in application-specific software, they cannot be used in
+combination for any arbitrary model structure.
+
+Using the open source modeling software NIMBLE, we develop combined
+computational approaches including user-defined and automatic blocking
+of parameters for MCMC, filtering over latent states, and customized
+MCMC samplers for specific parameters to improve efficiency. We test
+these approaches using three representative occupancy models of
+varying levels of complexity including a single species model with
+spatial auto-correlation, a single species dynamic (multi-season)
+model, and a multi-species model. We also develop and implement
+methods for calculating calibrated predictive posterior $p$-values to
+assess model fit and cross validation for model selection within
+NIMBLE.
 
@@
   spatial auto-correlation}
-The first model is a single species, single season occupancy model
-accounting for spatial auto-correlation. We let $z_{i}$ denote the
-true occupancy of a species at site $i$.  We then let $x_{i,j}$
-indicate whether we detected ($x_{i,j}=1$) or did not detect
-($x_{i,j}=0$) that species in the $j^{\mathrm{th}}$ visit to site $i$.
-We assumed that occupancy at the $i^{\mathrm{th}}$ site is a Bernoulli
-random variable $z_{i} \sim \mathrm{Bern}(\psi_{i})$ with probability
-$\psi_{i}$.  We included the effect of an arbitrary covariate (e.g.,
-elevation) on site occupancy. To model the spatial auto-correlation in
-occupancy between sites, we assume the co-variance between sites $Y_i$
-and $Y_j$ is a function of distance between $p_i$ and $p_j$. We
-computed the probability of occupancy at site $i$
+The first model we explore is a single species, single season
+occupancy model accounting for spatial auto-correlation. We let
+$z_{i}$ denote the true occupancy of a species at site $i$.  We then
+let $x_{i,j}$ indicate whether we detected ($x_{i,j}=1$) or did not
+detect ($x_{i,j}=0$) that species in the $j^{\mathrm{th}}$ visit to
+site $i$.  We assumed that occupancy at the $i^{\mathrm{th}}$ site is
+a Bernoulli random variable $z_{i} \sim \mathrm{Bern}(\psi_{i})$ with
+probability $\psi_{i}$.  We included the effect of an arbitrary
+covariate (e.g., elevation) on site occupancy. To model the spatial
+auto-correlation in occupancy between sites, we assume the co-variance
+between sites $Y_i$ and $Y_j$ is a function of distance between $p_i$
+and $p_j$. We computed the probability of occupancy at site $i$
 %
@@
 Where $\lambda$ is the exponential decay constant and $\sigma^2$ is
-SOMETHING....
+\flagged{SOMETHING...}
 
-To improve efficiency of this model...
+We simulate data for this model and then fit it using the default
+settings for NIMBLE and JAGS. \flagged{To improve efficiency of this
+  model...}
 
@@
 
-The second model is a relatively simple single species occupancy model
-over multiple seasons \citep{Royle2007}. We let $z_{i,t}$ denote the
-true occupancy of a species in year $t$ at site $i$.  We then let
-$x_{i,t,j}$ indicate whether we detected ($x_{i,t,j}=1$) or did not
-detect ($x_{i,t,j}=0$) that species in the $j^{\mathrm{th}}$ visit to
-site $i$ in year $t$.  We assumed that occupancy at the
-$i^{\mathrm{th}}$ site in the $t^{\mathrm{th}}$ year is a Bernoulli
-random variable $z_{i,t} \sim \mathrm{Bern}(\psi_{i,t})$ with
-probability $\psi_{i,t}$.
+The second model we examine is a relatively simple single species
+occupancy model over multiple seasons \citep{Royle2007}. We let
+$z_{i,j}$ denote the true occupancy of a species in year $j$ at site
+$i$.  We assumed that occupancy at the $i^{\mathrm{th}}$ site in the
+$j^{\mathrm{th}}$ year is a Bernoulli random variable $z_{i,j} \sim
+\mathrm{Bern}(\psi_{i,j})$.
 
-Letting $\phi_{i,t}$ denote the probability the species persists at
-site $i$ from years $t$ to $t+1$ (provided it was present at site $j$
-in year $t$, $z_{i,t}=1$) and $\gamma_{i,t}$ denote the
-probability that site $i$ is colonized in year $t+1$
-(provided it was not present at site $i$ in year $t$, $z_{i,t}=0$),
-we then computed the probability of occupancy  at site
-$i$ in subsequent years as
+Letting $\phi_{i,j}$ denote the probability the species persists at
+site $i$ from years $j$ to $j+1$ (provided it was present at site $i$
+in year $j$, $z_{i,j}=1$) and $\gamma_{i,j}$ denote the probability
+that site $i$ is colonized in year $j+1$ (provided it was not present
+at site $i$ in year $j$, $z_{i,j}=0$), we then computed the
+probability of occupancy at site $i$ in subsequent years as
 %
 \begin{equation}
-  \label{eq:occupancy2}
-  \psi_{i,t+1} =
-  \phi_{i,t} * z_{i,t} + \gamma_{i,t} * (1-z_{i,t})\hspace{0.2em}.
+  \label{eq:mutliseason}
+  \psi_{i,j+1} =
+  \phi_{i,j} * z_{i,k} + \gamma_{i,j} * (1-z_{i,j})\hspace{0.2em}.
 \end{equation}
@@
 
-First, to improve efficiency, filter over latent states
+We then let $x_{i,j,k}$ indicate whether we detected ($x_{i,j,k}=1$)
+or did not detect ($x_{i,j,k}=0$) that species in the
+$k^{\mathrm{th}}$ visit to site $i$ in year $j$. We assume detection
+was distributed according to be a Bernoulli random variable such that
+$x_{i,j,k} \sim \mathrm{Bern}(p_{j}*z_{i,j})$, where $p_{j}$ is the
+probability that the species was detected at site $i$ in the
+$j^{\mathrm{th}}$ year, given that it was present.
+
+As with the spatial occupancy model, we first simulate data for this
+model and then fit it using the default settings for JAGS and NIMBLE
+where all model parameters and latent states undergo MCMC sampling
+(``NIMBLE-latent'' and ``JAGS-latent'', respectively). Following
+\cite{Royle2007, kery-schaub-11}, we use uninformative
+$\mathrm{Unif}(0,1)$ priors for all parameters.
+
+Next, to improve efficiency, using NIMBLE we filter over latent states
 to calculate model likelihoods in order to limit MCMC sampling to
-top-level parameters. We then use two computational approaches to
-improve the efficiency of this model 1) dynamic blocking of the
-parameters \citep{turek2016efficient}, and 2) a custom MCMC
-specification where slice samplers \cite{neal-03} are used for all parameters.
+top-level parameters (``filter''). We then use two additional
+computational approaches to improve the efficiency of this model 1)
+dynamic blocking of the parameters \citep[``filter + blocking'',
+][]{turek2016efficient}, and 2) a custom MCMC specification where
+slice samplers \citep{neal-03} are used for all parameters (``filter +
+slice''). \flagged{other options we want to present?}
 
-Slice samplers are a class of methods that sample from a target distribution  by using that fact that samples from any distribution can be obtained by sampling uniformly from the area under that distribution's probability density function curve.
-The horizontal coordinates of these uniform samples will provide samples from the distribution of interest.
-Slice samplers have been shown to perform well in situations where choosing a tuning parameter for a Metropolis algorithm is difficult.
-
-%probably TMI below
-When used to sample from the posterior distribution of a univariate parameter, a slice sampler proceeds at each iteration by first choosing a vertical coordinate sampled uniformly between 0 and the height of the density curve at the parameter value from the previous iteration.
-Then, a horizontal coordinate is chosen uniformly from the set of all possible parameter values whose density is at least as great as the chosen vertical coordinate.
+Slice samplers are a class of methods that sample from a target
+distribution by using that fact that samples from any distribution can
+be obtained by sampling uniformly from the area under that
+distribution's probability density function curve.  The horizontal
+coordinates of these uniform samples will provide samples from the
+distribution of interest. Slice samplers have been shown to perform
+well in situations where choosing a tuning parameter for a Metropolis
+algorithm is difficult. When used to sample from the posterior
+distribution of a univariate parameter, a slice sampler proceeds at
+each iteration by first choosing a vertical coordinate sampled
+uniformly between 0 and the height of the density curve at the
+parameter value from the previous iteration.  Then, a horizontal
+coordinate is chosen uniformly from the set of all possible parameter
+values whose density is at least as great as the chosen vertical
+coordinate.
 
@@
 
-The last model is from \citep{zipkin2010multi}. It is a multi-species,
-single season occupancy model examining the effect of wildlife
-management and habitat characteristics on bird communities
-\citep{zipkin2010multi}. The species-specific coefficients for the
-effect of basal tree area, understory foliage and deer management
-where bound together by a common distribution with an estimated
-variance. The model is similar to Eq.~\ref{eq:spatial}, except each for
-species $k$, we let $z_{i,k}$ denote its true occupancy state at site
-$i$.
+The last model we analyze is a multi-species, single season occupancy
+model examining the effect of wildlife management and habitat
+characteristics on bird communities \citep{zipkin2010multi}. The
+species-specific coefficients for the effect of basal tree area,
+understory foliage and deer management where bound together by a
+common distribution with an estimated variance.  For species $i$, we
+let $z_{i,j}$ denote its true occupancy state at site $j$.  We then
+let $x_{i,j,k}$ indicate whether we detected ($x_{i,j,k}=1$) or did
+not detect ($x_{i,j,k}=0$) that species in the $k^{\mathrm{th}}$ visit
+to site $j$.  We assumed that the occupancy of the $i^{\mathrm{th}}$
+species at the $j^{\mathrm{th}}$ site is a Bernoulli random variable
+$z_{i,j} \sim \mathrm{Bern}(\psi_{i,j})$. We then let $x_{i,j,k}$
+indicate whether we detected ($x_{i,j,k}=1$) or did not detect
+($x_{i,j,k}=0$) species $i$ in the $k^{\mathrm{th}}$ visit to site
+$j$. We also assumed that detection was distributed according to be a
+Bernoulli random variable such that $x_{i,j,k} \sim
+\mathrm{Bern}(p_{i}*z_{i,j})$, where $p_{i}$ is the probability that
+the $i^{\mathrm{th}}$ species was detected. Both site occupancy and
+detection were influence by habiat and survey characteristics
+\citep{zipkin2010multi}. Specifically, occurence depended on the study
+area (CATO, Ind=1, or FCW, Ind=0), the basal tree area (BA) and the
+understory foliage cover (UFC). The species-specific occupancy
+probabilities are modeled as
+
+\begin{equation}
+  \begin{split}
+    \label{eq:multispecies}
+    \mathrm{logit}(\psi_{i,j}) &= uCATO_i(Ind_j) + uFCW_i(1-Ind_j) +
+    \alpha1_iUFC_j + \alpha2_iUFC^2_j + \alpha3_iBA_j +
+    \alpha4_iBA^2_j \hspace{0.2em}
+  \end{split}
+\end{equation}
+
+Similarly, detection survey location as well as the date and time
+since sunrise.
+
+\begin{equation}
+  \begin{split}
+    \label{eq:multispecies_detection}
+    \mathrm{logit}(p_{i,j,k}) &= vCATO_i(Ind_j) + vFCW_i(1-Ind_j) +
+    \beta1_idate_j + \beta2_idate^2_j + \beta3_isunrise_j
+    \hspace{0.2em}
+  \end{split}
+\end{equation}
 
@@
 
-Random walk Metropolis algorithms \citep{metropolis1953equation} sample from the posterior distribution of a parameter of interest by first proposing a new parameter value from a normal proposal distribution centered at the current value, and then deciding whether to accept or reject the proposed value via the Metropolis ratio.
-Although common, standard univariate or multivariate normal proposal distributions can prove to be inefficient at sampling for models in which parameters are highly correlated.
-Adaptive random walk Metropolis sampling \citep{haario98anadaptive} allows the correlation structure of the posterior distribution of blocks of parameters to be used to produce better proposals.
-For an occupancy model where parameters are highly correlated within species, this can provide much more efficient sampling than a non-adaptive Metropolis algorithm.
+Random walk Metropolis algorithms \citep{metropolis1953equation}
+sample from the posterior distribution of a parameter of interest by
+first proposing a new parameter value from a normal proposal
+distribution centered at the current value, and then deciding whether
+to accept or reject the proposed value via the Metropolis ratio.
+Although common, standard univariate or multivariate normal proposal
+distributions can prove to be inefficient at sampling for models in
+which parameters are highly correlated.  Adaptive random walk
+Metropolis sampling \citep{haario98anadaptive} allows the correlation
+structure of the posterior distribution of blocks of parameters to be
+used to produce better proposals.  For an occupancy model where
+parameters are highly correlated within species, this can provide much
+more efficient sampling than a non-adaptive Metropolis algorithm.
 

### 2016-10-11 analysis/crossValidation/crossValidationFunction.R

@@ -27,7 +27,6 @@ crossValidateOne <- function(model,
   ## the average loss (0/1) can be taken over all MCMC runs.  then
   ## take the average of these for all data points? woo!
   newModel <- model$newModel()
-  compileNimble(newModel)
   numBlocks <- newModel$getVarInfo(dataNames)[['maxs']][leaveOutIndex]
   dataDimensions <- newModel$getVarInfo(dataNames)[['maxs']]
   saveData <- array(nimble:::values(newModel, dataNames),

### 2016-10-11 analysis/multiSpp-singleSea/opt1-4.R

@@ -288,7 +288,7 @@ occ.R.model <- nimbleModel(code=ms.ss.occ,
                            inits=input1$inits,
                            check=FALSE)
 
-options(mc.cores=6)
+options(mc.cores=1)
 
 source('../crossValidation/crossValidationFunction.R')
 

### 2016-10-11 analysis/singleSpp-multiSea/opt2-3.R

@@ -1,9 +1,8 @@
 rm(list=ls())
 setwd("~/Dropbox/nimble/occupancy/analysis/singleSpp-multiSea")
-## setwd("C:/Users/iateb/Dropbox/nimble/occupancy/analysis/singleSpp-multiSea")
 
 source('src/initialize.R')
-
+set.seed(444)
 data <- genDynamicOccData()
 model.input <- prepModDataOcc(data)
 
@@ -50,33 +49,33 @@ input1 <- c(code=ss.ms.occ,
 ## opt 2: add custom z sampler and slice on uniform(0,1) nodes
 ## *********************************************************************
 
-## MCMCdefs.opt2 <- list('nimbleOpt2' = quote({
-##   customSpec <- configureMCMC(Rmodel)
-##   customSpec$removeSamplers('phi', print=FALSE)
-##   customSpec$removeSamplers('gamma', print=FALSE)
-##   customSpec$removeSamplers('p', print=FALSE)
-##   customSpec$removeSamplers('psi1', print=FALSE)
-##   ## happens to be all top nodes
-##   zeroOneNodes <- Rmodel$getNodeNames(topOnly = TRUE)
-##   for(zon in zeroOneNodes) customSpec$addSampler(target = zon,
-##                                                  type =
-##                                                  "slice",
-##                                                  print=FALSE)
-##   customSpec
-## }))
+MCMCdefs.opt2 <- list('nimbleOpt2' = quote({
+  customSpec <- configureMCMC(Rmodel)
+  customSpec$removeSamplers('phi', print=FALSE)
+  customSpec$removeSamplers('gamma', print=FALSE)
+  customSpec$removeSamplers('p', print=FALSE)
+  customSpec$removeSamplers('psi1', print=FALSE)
+  ## happens to be all top nodes
+  zeroOneNodes <- Rmodel$getNodeNames(topOnly = TRUE)
+  for(zon in zeroOneNodes) customSpec$addSampler(target = zon,
+                                                 type =
+                                                 "slice",
+                                                 print=FALSE)
+  customSpec
+}))
 
 ## *********************************************************************
 ## run with compareMCMCs
 
-## ss.ms.opt2 <- compareMCMCs(input1,
-##                            MCMCs=c('nimbleOpt2'),
-##                            MCMCdefs = MCMCdefs.opt2,
-##                            niter= niter,
-##                            burnin = burnin,
-##                            summary=FALSE,
-##                            check=FALSE)
[…]

### 2016-10-11 analysis/singleSpp-multiSea/opt4-5.R

@@ -2,6 +2,7 @@ rm(list=ls())
 
 setwd("~/Dropbox/nimble/occupancy/analysis/singleSpp-multiSea")
 source('src/initialize.R')
+set.seed(444)
 data <- genDynamicOccData()
 model.input <- prepModDataOcc(data, include.zs=FALSE)
 
@@ -94,28 +95,28 @@ occ.R.model <- nimbleModel(code=ss.ms.occ,
                            data=input1$data,
                            inits=input1$inits,
                            check=FALSE)
-## ad auto blocking
+
 occ.mcmc <- buildMCMC(occ.R.model)
 occ.C.model <- compileNimble(occ.R.model)
 occ.C.mcmc <- compileNimble(occ.mcmc, project = occ.R.model)
-occ.C.mcmc$run(10000)
+occ.C.mcmc$run(niter)
 
 source('../cppp/src/calcCPPP.R', chdir = TRUE)
-options(mc.cores=6)
+options(mc.cores=1)
 
-test.opt2 <- generateCPPP(occ.R.model,
+test.opt4 <- generateCPPP(occ.R.model,
                           occ.C.model,
                           occ.C.mcmc,
                           occ.mcmc,
                           dataName = 'y',
                           paramNames = input1$monitors, 
-                          MCMCIter = 10000, 
-                          NSamp = 1000,
-                          NPDist = 100,
+                          MCMCIter = niter, 
+                          NSamp = 10^3,
+                          NPDist = 10^3,
                           burnInProportion = 0.10,
                           thin = 1,
                           averageParams = TRUE,
                           discFuncGenerator=likeDiscFuncGenerator)
 
-save(test.opt2, file=file.path(save.dir, "ssms_noz_CPPP.Rdata"))
+save(test.opt4, file=file.path(save.dir, "ssms_noz_CPPP.Rdata"))
 

### 2016-10-11 analysis/singleSpp-multiSea/original.R

@@ -2,6 +2,7 @@ rm(list=ls())
 setwd("~/Dropbox/nimble/occupancy/analysis/singleSpp-multiSea")
 source('src/initialize.R')
 
+set.seed(444)
 data <- genDynamicOccData()
 model.input <- prepModDataOcc(data)
 
@@ -60,7 +61,7 @@ input1 <- c(code=ss.ms.occ,
 
 
 ss.ms.orig <- compareMCMCs(input1,
-                           MCMCs=c('jags','nimble'),
+                           MCMCs=c('jags', "nimble"),
                            niter=niter,
                            burnin = burnin,
                            summary=FALSE,

### 2016-10-11 analysis/singleSpp-multiSea/src/dataGen.R

@@ -22,7 +22,7 @@ expit <- function(x) {
 ## range.psi and range.gamma - same for survival and colonization probability
 
 genDynamicOccData <- function(R = 250,
-                              J = 3, K = 10,
+                              J = 10, K = 10,
                               psi1 = 0.4,
                               range.p = c(0.2, 0.4),
                               range.phi = c(0.6, 0.8),

### 2016-10-11 analysis/singleSpp-multiSea/src/initialize.R

@@ -1,7 +1,7 @@
-## library(devtools)
-## install_github("nimble-dev/nimble",
-##                ref = "dlevel",
-##                subdir = "packages/nimble")
+library(devtools)
+install_github("nimble-dev/nimble",
+               ref = "devel",
+               subdir = "packages/nimble")
 
 library(nimble)
 library(igraph)

### 2016-10-11 analysis/spatial/opt1-3.R

@@ -5,7 +5,7 @@ source('src/initialize.R')
 set.seed(444)
 dats <- genSpatialOccData()
 model.input <- prepModData(dats$data, dats$y, dats$distance,
-                           nsite=50)
+                           nsite=100)
 
 sp.mod <- nimbleCode({
   ## priors
@@ -56,40 +56,40 @@ save(sp.opt1, file=file.path(save.dir, "opt1.Rdata"))
 ## *********************************************************************
 ## opt 2: add custom z sampler and slice on uniform(0,1) nodes
 ## *********************************************************************
-library(devtools)
-install_github("nlmichaud/nimble/packages/nimble", ref= "AFSliceSampler")
+## library(devtools)
+## install_github("nlmichaud/nimble/packages/nimble", ref= "AFSliceSampler")
 
-MCMCdefs.opt2 <- list('nimbleOpt2' = quote({
-  customSpec <- configureMCMC(Rmodel)
-  ## slice sampler for 0,1 varaibles
-  customSpec$removeSamplers('p', print=FALSE)
-  customSpec$addSampler(target = 'p',
-                        type =
-                        "slice",
-                        print=FALSE)
-  ## multivariate slice sampler
-  customSpec$removeSamplers('rho', print=FALSE)
-  customSpec$addSampler(target = 'rho',
-                                                 type =
-                                                   "AF_slice",
-                                                 print=FALSE,
-                        control = list(sliceWidths =
-                          rep(1, input1$constants$nsite),
-                          sliceFactorBurnInIters = 5000,
-                          sliceFactorAdaptInterval = 200,
-                          sliceSliceAdaptIters = 200))
-  customSpec
-}))
+## MCMCdefs.opt2 <- list('nimbleOpt2' = quote({
+##   customSpec <- configureMCMC(Rmodel)
+##   ## slice sampler for 0,1 varaibles
+##   customSpec$removeSamplers('p', print=FALSE)
+##   customSpec$addSampler(target = 'p',
+##                         type =
+##                         "slice",
+##                         print=FALSE)
+##   ## multivariate slice sampler
+##   customSpec$removeSamplers('rho', print=FALSE)
+##   customSpec$addSampler(target = 'rho',
+##                                                  type =
+##                                                    "AF_slice",
+##                                                  print=FALSE,
+##                         control = list(sliceWidths =
+##                           rep(1, input1$constants$nsite),
+##                           sliceFactorBurnInIters = 5000,
[…]

### 2016-10-11 analysis/spatial/original.R

@@ -7,7 +7,7 @@ load.module("msm")
 set.seed(444)
 dats <- genSpatialOccData()
 model.input <- prepModData(dats$data, dats$y, dats$distance,
-                           nsite=50)
+                           nsite=100)
 
 mexp <- nimbleFunction(
   run = function(A = double(2)){
@@ -46,15 +46,14 @@ sp.mod <- nimbleCode({
   ## JAGS)
 
   ## mexp is jags's version fo matrix exponentiation, very sensitive 
-  ## temp.cov[1:nsite, 1:nsite] <- -delta*D[1:nsite, 1:nsite]
-  D.cov[1:nsite, 1:nsite]  <- (sigma^2)*mexp(-delta*D[1:nsite, 1:nsite])
+  ## D.cov[1:nsite, 1:nsite]  <- (sigma^2)*mexp(-delta*D[1:nsite, 1:nsite])
 
-  ## for(i in 1:nsite){
-  ##   for(j in 1:nsite){
-  ##     temp.cov[i, j] <- -delta*D[i, j]
-  ##     D.cov[i, j]  <- (sigma^2)* exp(temp.cov[i, j])
-  ##   }
-  ## }
+  for(i in 1:nsite){
+    for(j in 1:nsite){
+      temp.cov[i, j] <- -delta*D[i, j]
+      D.cov[i, j]  <- (sigma^2)*exp(temp.cov[i, j])
+    }
+  }
   
   D.tau[1:nsite, 1:nsite] <- inverse(D.cov[1:nsite, 1:nsite])
   

### 2016-10-11 analysis/spatial/src/dataGen.R

@@ -10,10 +10,10 @@ rmvn <- function(n, mu = 0, V = matrix(1)) {
 
 
 genSpatialOccData <- function(ngrid = 50,
-                              nreps = 10,
+                              nreps = 50,
                               alpha = 1,
-                              beta1 = 2,
-                              p = 0.6,
+                              beta1 = 6,
+                              p = 0.8,
                               sigma = 0.5,
                               delta = 0.5){
 
@@ -36,7 +36,7 @@ genSpatialOccData <- function(ngrid = 50,
 
   ## calculate probabilities of occurrence
   psi <- expit(alpha + beta1 * raster::values(elev) +
-               raster::values(Xraster))
+               raster::values(Xraster))  
 
   ## Latent occurrence state
   z <- rbinom(n = n, size = 1, prob = psi) 

### 2016-10-18 ms/ms/ms.tex
misc bug fixes

@@
 to calculate model likelihoods in order to limit MCMC sampling to
-top-level parameters (``filter''). We then use two additional
-computational approaches to improve the efficiency of this model 1)
-dynamic blocking of the parameters \citep[``filter + blocking'',
+top-level parameters (``filter''). \flagged{Do we want to write out
+  the likelihood?)} We then use two additional computational
+approaches to improve the efficiency of this model 1) dynamic blocking
+of the parameters \citep[``filter + autoblocking'',
 ][]{turek2016efficient}, and 2) a custom MCMC specification where
 slice samplers \citep{neal-03} are used for all parameters (``filter +
-slice''). \flagged{other options we want to present?}
-
-Slice samplers are a class of methods that sample from a target
-distribution by using that fact that samples from any distribution can
-be obtained by sampling uniformly from the area under that
-distribution's probability density function curve.  The horizontal
-coordinates of these uniform samples will provide samples from the
-distribution of interest. Slice samplers have been shown to perform
-well in situations where choosing a tuning parameter for a Metropolis
-algorithm is difficult. When used to sample from the posterior
-distribution of a univariate parameter, a slice sampler proceeds at
-each iteration by first choosing a vertical coordinate sampled
-uniformly between 0 and the height of the density curve at the
+slice''). Slice samplers are a class of methods that sample from a
+target distribution by using that fact that samples from any
+distribution can be obtained by sampling uniformly from the area under
+that distribution's probability density function curve.  The
+horizontal coordinates of these uniform samples will provide samples
+from the distribution of interest. Slice samplers have been shown to
+perform well in situations where choosing a tuning parameter for a
+Metropolis algorithm is difficult. When used to sample from the
+posterior distribution of a univariate parameter, a slice sampler
+proceeds at each iteration by first choosing a vertical coordinate
+sampled uniformly between 0 and the height of the density curve at the
 parameter value from the previous iteration.  Then, a horizontal
@@
 values whose density is at least as great as the chosen vertical
-coordinate.
+coordinate.  \flagged{other options we want to present?}
 
@@
 
+We first fit the using the default settings for JAGS and NIMBLE where
+all model parameters and latent states undergo MCMC sampling
+(``NIMBLE-latent'' and ``JAGS-latent'', respectively). We use
+uninformative priors $\mathrm{Norm}(0,1000)$ for the means of the
+distributions of the hyperparameters and $\mathrm{Unif}(0,100)$ the
+variances.
+
 To improve the efficiency of this model, we first filtered over latent
 states to calculate model likelihoods in order to limit MCMC sampling
-to top-level parameters. We also vectorized all calculations that
-would have require for loops in BUGS or JAGS. We then tried two
-approaches to speed sampling of the top-level parameters 1) dynamic
-blocking of the parameters \citep{turek2016efficient}, and 2) a custom
-blocking scheme where the parameters of each species are blocked
-together with adaptive random walk MCMC.
-
-Random walk Metropolis algorithms \citep{metropolis1953equation}
-sample from the posterior distribution of a parameter of interest by
-first proposing a new parameter value from a normal proposal
-distribution centered at the current value, and then deciding whether
-to accept or reject the proposed value via the Metropolis ratio.
-Although common, standard univariate or multivariate normal proposal
-distributions can prove to be inefficient at sampling for models in
-which parameters are highly correlated.  Adaptive random walk
-Metropolis sampling \citep{haario98anadaptive} allows the correlation
-structure of the posterior distribution of blocks of parameters to be
-used to produce better proposals.  For an occupancy model where
-parameters are highly correlated within species, this can provide much
-more efficient sampling than a non-adaptive Metropolis algorithm.
+to top-level parameters (``Filter''). We also vectorized all
+calculations that would have require for loops in JAGS. \flagged{Do we
+  want to write out the likelihood?)} We then applied two approaches
+to speed sampling of the top-level parameters 1) dynamic blocking of
+the parameters \citep[``Filter + autoblocking'',
+][]{turek2016efficient}, and 2) a custom blocking scheme where the
+parameters of each species are blocked together with adaptive random
+walk MCMC (``Filter + species blocking''). Random walk Metropolis
+algorithms \citep{metropolis1953equation} sample from the posterior
+distribution of a parameter of interest by first proposing a new
+parameter value from a normal proposal distribution centered at the
+current value, and then deciding whether to accept or reject the
+proposed value via the Metropolis ratio.  Although common, standard
+univariate or multivariate normal proposal distributions can prove to
+be inefficient at sampling for models in which parameters are highly
+correlated.  Adaptive random walk Metropolis sampling
+\citep{haario98anadaptive} allows the correlation structure of the
+posterior distribution of blocks of parameters to be used to produce
+better proposals.  For an occupancy model where parameters are highly
+correlated within species, this can provide much more efficient
+sampling than a non-adaptive Metropolis algorithm.
 
@@
 fitting, then use the fitted model to predict $y_k$. The prediction
-error is summarized by comparing the simulated $y_k$ to the true $y_k$.
+error is summarized by comparing the simulated $y_k$ to the true
+$y_k$.
 
-In the multi species occupancy models (Section \ref{sec:msss}),
-practitioners are often interested in determining whether a model
-including a species random effect for explanatory variables is a
-better fit than a model without the random effect. We implemented a
-cross-validation procudure for this model where the detection data for
-species is left out, the model refitted, and the fitted model used to
-predict the occurence of that species. The predictive error of the
-model included a random effect of species is then compared to a model
-where no species random effects were included.
+In multi-species occupancy models like the model we explored in
+Section \ref{sec:msss}), practitioners are often interested in
+determining whether a model including a species random effect for
+explanatory variables is a better fit than a model without the random
+effect. We implemented a cross-validation procedure for this model
+where the detection data for species is left out, the model refitted,
+and the fitted model used to predict the occurence of that
+species. The predictive error of the model included a random effect of
+species is then compared to a model where no species random effects
+were included.
 

### 2016-10-18 analysis/all/plotting.R

@@ -1,10 +1,11 @@
+library(RColorBrewer)
 
-  pdf.f <- function(f, file, ...) {
-    cat(sprintf('Writing %s\n', file))
-    pdf(file, ...)
-    on.exit(dev.off())
-    f()
-  }
+pdf.f <- function(f, file, ...) {
+  cat(sprintf('Writing %s\n', file))
+  pdf(file, ...)
+  on.exit(dev.off())
+  f()
+}
 
 ## function to plot values of parameters as a function of
 ## iterations. Input is a MCMCcompare object
@@ -32,4 +33,40 @@ checkChains <- function(all.mods.samps, f.path){
           file= file.path(sprintf(f.path, z)),
           height=11, width=8.5)
   })
+}
+
+
+plotEffSize <- function(eff.size, eff.param,
+                        f.path, name, at, adj=0.5){
+  f <- function(){
+    layout(matrix(1:4, ncol=2))
+    par(oma=c(5, 7.5, 0.5, 1),
+        mar=c(0.5, 0, 3, 1), cex.axis=1.5)
+    cols <- brewer.pal(length(eff.size$mean), "Greys")
+    ## barplots
+    mp1 <- barplot(eff.size$mean, names="", las=1, col=cols)
+    mtext("Mean", 3, line=0.5, cex=1.5)
+
+    mp2 <- barplot(eff.size$min, names="", las=1, col=cols)
+    text(mp2, par('usr')[3] - adj,
+         srt = 45, adj = 1,
+         labels = names(eff.size$mean),
+         xpd = NA,
+         cex=1)
+
+    mtext("Minimum", 3, line=0.5, cex=1.5)
+    mtext("Effective sample size \n per second",
+          2, line=4.5, cex=1.5, at=at)
+    ## for each parm
+    plot(NA, ylim=range(eff.param), xlim=c(1, ncol(eff.param)))
+    for(i in 1:nrow(eff.param)){
+      points(x=1:ncol(eff.param), y=eff.param[i,], col=cols[i],
+             pch=16,
+             xaxt="n")
+    }
+    
+  }
+  pdf.f(f,
+        file= file.path(sprintf(f.path, name)),
+        height=6, width=4.5)
 }

### 2016-10-18 analysis/cppp/src/calcCPPP.R

@@ -138,7 +138,7 @@ generateCPPP <-  function(R.model,
                        NSamp,
                        C.pppFunc,
                        orig.C.mcmc,
-                       firstRun = 1)
+                       firstRun = 1)$pre.pp
 
   ## refits model with sampled data, reruns, enter inner loop,
   ## calculates distbution of PPPs

### 2016-10-18 analysis/multiSpp-singleSea/compare.R

@@ -1,7 +1,8 @@
 rm(list=ls())
-## gctorture()
 
 library(nimble)
+library(coda)
+
 setwd('~/Dropbox/nimble/occupancy/analysis/multiSpp-singleSea')
 source('../all/plotting.R')
 save.dir <-  "../../../saved/multiSpp-singleSea/saved"
@@ -24,9 +25,20 @@ load(file=file.path(save.dir, "opt2.Rdata"))
 ## option 1 + sigma sampler on random effects
 load(file=file.path(save.dir, "opt4.Rdata"))
 
-ms.ss.opt1[[1]] <- rename_MCMC_comparison_method('nimble', 'remove_z',
+
+
+ms.ss.orig[[1]] <- rename_MCMC_comparison_method(c('nimble', 'jags'),
+                                                 c('NIMBLE-latent',
+                                                   'JAGS-latent'),
+                                                 comparison=ms.ss.orig[[1]])
+
+ms.ss.opt1[[1]] <- rename_MCMC_comparison_method(c('nimble',
+                                                   "autoBlock"),
+                                                 c('filter',
+                                                   'filter + autoblock'),
                                                  comparison=ms.ss.opt1[[1]])
-ms.ss.opt2[[1]] <- rename_MCMC_comparison_method('nimbleOpt2', 'block_1',
+ms.ss.opt2[[1]] <- rename_MCMC_comparison_method('nimbleOpt2',
+                                                 'filter + sp. block',
                                                  comparison=ms.ss.opt2[[1]])
 ## ms.ss.opt3[[1]] <- rename_MCMC_comparison_method('nimbleOpt3',
 ##                                                  'block_2',
@@ -40,7 +52,7 @@ ms.ss.occ.all <- combine_MCMC_comparison_results(ms.ss.orig[[1]],
                                                  ms.ss.opt1[[1]],
                                                  ms.ss.opt2[[1]],
                                                  ## ms.ss.opt3[[1]],
-                                                 ms.ss.opt4[[1]],
+                                                 ## ms.ss.opt4[[1]],
                                                  name = "ms.ss" )
 
 make_MCMC_comparison_pages(ms.ss.occ.all,
@@ -53,3 +65,15 @@ checkChains(ms.ss.occ.all[[1]]$samples,
             f.path = file.path(save.dir,
             "../figures/chains/%s.pdf")
 )
+
+
+## ****************************************
[…]

### 2016-10-18 analysis/multiSpp-singleSea/opt1-4.R

@@ -258,7 +258,7 @@ occ.C.mcmc <- compileNimble(occ.mcmc, project = occ.R.model)
 occ.C.mcmc$run(niter)
 
 source('../cppp/src/calcCPPP.R', chdir = TRUE)
-options(mc.cores=6)
+options(mc.cores=8)
 
 test.opt2 <- generateCPPP(occ.R.model,
                           occ.C.model,

### 2016-10-18 analysis/singleSpp-multiSea/compare.R

@@ -19,6 +19,12 @@ load(file=file.path(save.dir, "opt5.Rdata"))
 
 ## rename results
 
+ss.ms.orig[[1]] <- rename_MCMC_comparison_method(c('nimble', 'jags'),
+                                                 c('NIMBLE-latent',
+                                                   'JAGS-latent'),
+                                                 comparison=ss.ms.orig[[1]])
+
+
 ss.ms.opt2[[1]] <- rename_MCMC_comparison_method('nimbleOpt2',
                                                  'slice',
                                                  comparison=ss.ms.opt2[[1]])
@@ -26,19 +32,19 @@ ss.ms.opt2[[1]] <- rename_MCMC_comparison_method('nimbleOpt2',
 ss.ms.opt4[[1]] <- rename_MCMC_comparison_method(c('nimble',
                                                    'autoBlock',
                                                    'nimble_slice'),
-                                                 c('no z',
-                                                   'no z, autoBlock',
-                                                   'no z, slice'),
+                                                 c('filter',
+                                                   'filter + autoblock',
+                                                   'filter + slice'),
                                                  comparison=ss.ms.opt4[[1]])
 
-ss.ms.opt5[[1]] <- rename_MCMC_comparison_method('nimbleOpt5',
-                                                 'block phi gam',
-                                                 comparison=ss.ms.opt5[[1]])
+## ss.ms.opt5[[1]] <- rename_MCMC_comparison_method('nimbleOpt5',
+##                                                  'block phi gam',
+##                                                  comparison=ss.ms.opt5[[1]])
 ## compare mcmcs
 ss.ms.occ.all <- combine_MCMC_comparison_results(ss.ms.orig[[1]],
-                                                 ss.ms.opt2[[1]],
+                                                 ## ss.ms.opt2[[1]],
                                                  ss.ms.opt4[[1]],
-                                                 ss.ms.opt5[[1]],
+                                                 ## ss.ms.opt5[[1]],
                                                  name = "ss.ms" )
 
 make_MCMC_comparison_pages(ss.ms.occ.all,
@@ -47,5 +53,16 @@ make_MCMC_comparison_pages(ss.ms.occ.all,
 
 checkChains(ss.ms.occ.all[[1]]$samples,
             f.path = file.path(save.dir,
-            "../figures/chains/%s.pdf")
-)
+              "../figures/chains/%s.pdf")
+            )
+
[…]

### 2016-10-18 analysis/spatial/opt1-3.R

@@ -5,11 +5,11 @@ source('src/initialize.R')
 set.seed(444)
 dats <- genSpatialOccData()
 model.input <- prepModData(dats$data, dats$y, dats$distance,
-                           nsite=100)
+                           nsite=250)
 
 sp.mod <- nimbleCode({
   ## priors
-  delta ~ dunif(0.01, 10)
+  delta ~ dunif(0.1, 10)
   sigma ~ dunif(0, 10)
   p ~ dunif(0, 1)
   alpha ~ dnorm(0, 0.001)

### 2016-10-18 analysis/spatial/original.R

@@ -7,7 +7,7 @@ load.module("msm")
 set.seed(444)
 dats <- genSpatialOccData()
 model.input <- prepModData(dats$data, dats$y, dats$distance,
-                           nsite=100)
+                           nsite=250)
 
 mexp <- nimbleFunction(
   run = function(A = double(2)){
@@ -18,7 +18,7 @@ mexp <- nimbleFunction(
 
 sp.mod <- nimbleCode({
   ## priors
-  delta ~ dunif(0.01, 10)
+  delta ~ dunif(0.1, 10)
   sigma ~ dunif(0, 10)
   p ~ dunif(0, 1)
   alpha ~ dnorm(0, 0.001)

### 2016-10-18 analysis/spatial/src/dataGen.R

@@ -80,7 +80,9 @@ prepModData <- function(fulldata, y, distance, nsite,
   zinits[zinits == 1] <- 2
   zinits[is.na(zinits)] <- 1
   zinits[zinits == 2] <- NA
-  inits <- list(z = zinits)
+  inits <- list(z = zinits,
+                sigma=0.1,
+                delta=0.1)
 
   model.data <- list(D = distance[sites, sites],
                      y = y,

### 2016-10-18 analysis/spatial/src/initialize.R

@@ -1,7 +1,7 @@
-## library(devtools)
-## install_github("nimble-dev/nimble",
-##                ref = "devel",
-##                subdir = "packages/nimble")
+library(devtools)
+install_github("nimble-dev/nimble",
+               ref = "devel",
+               subdir = "packages/nimble")
 
 library(nimble)
 library(igraph)

### 2016-10-18 analysis/all/plotting.R
made new plotting funciton to plot efficiency by parameter

@@ -37,18 +37,18 @@ checkChains <- function(all.mods.samps, f.path){
 
 
 plotEffSize <- function(eff.size, eff.param,
-                        f.path, name, at, adj=0.5){
+                        f.path, name, at, adj1, adj2){
+  cols <- brewer.pal(length(eff.size$mean)+1, "Greys")[-1]
   f <- function(){
-    layout(matrix(1:4, ncol=2))
+    layout(matrix(1:2, ncol=1))
     par(oma=c(5, 7.5, 0.5, 1),
         mar=c(0.5, 0, 3, 1), cex.axis=1.5)
-    cols <- brewer.pal(length(eff.size$mean), "Greys")
     ## barplots
     mp1 <- barplot(eff.size$mean, names="", las=1, col=cols)
     mtext("Mean", 3, line=0.5, cex=1.5)
 
     mp2 <- barplot(eff.size$min, names="", las=1, col=cols)
-    text(mp2, par('usr')[3] - adj,
+    text(mp2, par('usr')[3] - adj1,
          srt = 45, adj = 1,
          labels = names(eff.size$mean),
          xpd = NA,
@@ -57,16 +57,40 @@ plotEffSize <- function(eff.size, eff.param,
     mtext("Minimum", 3, line=0.5, cex=1.5)
     mtext("Effective sample size \n per second",
           2, line=4.5, cex=1.5, at=at)
-    ## for each parm
-    plot(NA, ylim=range(eff.param), xlim=c(1, ncol(eff.param)))
-    for(i in 1:nrow(eff.param)){
-      points(x=1:ncol(eff.param), y=eff.param[i,], col=cols[i],
-             pch=16,
-             xaxt="n")
-    }
+  }
+
+  f2 <- function(){
+    layout(matrix(1:2, ncol=1), heights=c(1,3))
+
+    par(oma=c(5, 6, 0.5, 1),
+        mar=c(0.5, 0, 0.5, 1), cex.axis=1.5)
+
+    plot(NA, ylim=c(0,1), xlim=c(0,1), yaxt= "n", xaxt="n", bty="n")
+    legend("top", legend=rownames(eff.param),
+           pch=16, col=cols, bty="n", ncol=3)
     
+    plot(NA, ylim=log(range(eff.param)), xlim=c(1, ncol(eff.param)),
+         xlab="", ylab="", xaxt="n")
+    for(i in 1:nrow(eff.param)){
+      points(x=1:ncol(eff.param), y=log(eff.param[i,]),
+             col=cols[i],
[…]

### 2016-10-18 analysis/multiSpp-singleSea/compare.R

@@ -53,7 +53,7 @@ ms.ss.occ.all <- combine_MCMC_comparison_results(ms.ss.orig[[1]],
                                                  ms.ss.opt2[[1]],
                                                  ## ms.ss.opt3[[1]],
                                                  ## ms.ss.opt4[[1]],
-                                                 name = "ms.ss" )
+                                                 name = "ms.ss"))
 
 make_MCMC_comparison_pages(ms.ss.occ.all,
                            dir=file.path(save.dir, "../figures/comparisons"))
@@ -74,6 +74,8 @@ checkChains(ms.ss.occ.all[[1]]$samples,
 by.param <- apply(ms.ss.occ.all[[1]]$samples, c(1,2), effectiveSize)
 by.config <- ms.ss.occ.all[[1]]$efficiency
 
+
+source('../all/plotting.R')
 plotEffSize(by.config, by.param, f.path= file.path(save.dir,
-              "../figures/comparisons/%s.pdf"), "MultiSpp-SingleSea",
-            at=0.4, adj=0.03)
+              "../figures/comparisons/%s%s.pdf"), "MultiSpp-SingleSea",
+            at=0.4, adj1=0.03, adj2=0.1)

### 2016-10-18 analysis/all/plotting.R
updated efficieincy plots

@@ -37,8 +37,10 @@ checkChains <- function(all.mods.samps, f.path){
 
 
 plotEffSize <- function(eff.size, eff.param,
-                        f.path, name, at, adj1, adj2){
+                        f.path, name, at, adj1, adj2,
+                        widths=c(4.5, 6)){
   cols <- brewer.pal(length(eff.size$mean)+1, "Greys")[-1]
+  names(cols) <- names(eff.size$mean)
   f <- function(){
     layout(matrix(1:2, ncol=1))
     par(oma=c(5, 7.5, 0.5, 1),
@@ -82,15 +84,15 @@ plotEffSize <- function(eff.size, eff.param,
          srt = 45, adj = 1 + adj2,
          labels = colnames(eff.param),
          xpd = NA,
-         cex=1)
+         cex=0.8)
     mtext("Effective sample size \n per second (log)",
           2, line=3.2, cex=1.5)
   }
   pdf.f(f,
         file= file.path(sprintf(f.path, name, "Bar")),
-        height=6, width=4.5)
+        height=6, width=widths[1])
   pdf.f(f2,
         file= file.path(sprintf(f.path, name, "Points")),
-        height=4, width=6)
+        height=4, width=widths[2])
 }
 

### 2016-10-18 analysis/multiSpp-singleSea/compare.R

@@ -53,7 +53,7 @@ ms.ss.occ.all <- combine_MCMC_comparison_results(ms.ss.orig[[1]],
                                                  ms.ss.opt2[[1]],
                                                  ## ms.ss.opt3[[1]],
                                                  ## ms.ss.opt4[[1]],
-                                                 name = "ms.ss"))
+                                                 name = "ms.ss")
 
 make_MCMC_comparison_pages(ms.ss.occ.all,
                            dir=file.path(save.dir, "../figures/comparisons"))
@@ -61,17 +61,18 @@ make_MCMC_comparison_pages(ms.ss.occ.all,
 
 ## look at samples
 
-checkChains(ms.ss.occ.all[[1]]$samples,
-            f.path = file.path(save.dir,
-            "../figures/chains/%s.pdf")
-)
+## checkChains(ms.ss.occ.all[[1]]$samples,
+##             f.path = file.path(save.dir,
+##             "../figures/chains/%s.pdf")
+## )
 
 
 ## ****************************************
 ## custom figs
 ## ****************************************
 
-by.param <- apply(ms.ss.occ.all[[1]]$samples, c(1,2), effectiveSize)
+by.param <- apply(ms.ss.occ.all[[1]]$samples, c(1,2), effectiveSize)/
+  ms.ss.occ.all[[1]]$timing 
 by.config <- ms.ss.occ.all[[1]]$efficiency
 
 

### 2016-10-18 analysis/singleSpp-multiSea/compare.R

@@ -51,18 +51,22 @@ make_MCMC_comparison_pages(ss.ms.occ.all,
                            dir=file.path(save.dir, "../figures/comparisons"))
 
 
-checkChains(ss.ms.occ.all[[1]]$samples,
-            f.path = file.path(save.dir,
-              "../figures/chains/%s.pdf")
-            )
+## checkChains(ss.ms.occ.all[[1]]$samples,
+##             f.path = file.path(save.dir,
+##               "../figures/chains/%s.pdf")
+##             )
+
 
 ## ****************************************
 ## custom figs
 ## ****************************************
 
-by.param <- apply(ss.ms.occ.all[[1]]$samples, c(1,2), effectiveSize)
+by.param <- apply(ss.ms.occ.all[[1]]$samples, c(1,2), effectiveSize)/
+  ss.ms.occ.all[[1]]$timing
 by.config <- ss.ms.occ.all[[1]]$efficiency
 
-plotEffSize(by.param, by.config, f.path= file.path(save.dir,
-              "../figures/comparisons/%s.pdf"), name="singleSpp-MultiSea",
-            at=9)
+
+source('../all/plotting.R')
+plotEffSize(by.config, by.param, f.path= file.path(save.dir,
+              "../figures/comparisons/%s%s.pdf"), "SingleSpp-MultiSea",
+            at=9, adj1=0.3, adj2=0.3, widths=c(4.5, 8.5))

### 2016-10-18 analysis/singleSpp-multiSea/occEx.R

@@ -1,242 +1,0 @@
-library(nimble)
-library(igraph)
-
-distributionsInputList <- nimble:::distributionsInputList
-## OR
-## library(devtools)
-## install_github("lponisio/nimble",
-##                ref = "userDistWorkArnd",
-##                subdir = "packages/nimble")
-
-## *********************************************************************
-## user defined functions to remove latent states
-## *********************************************************************
-
-## DynamicOccupancy removes the z's and muZ's from the model and computes
-## the probability of all reps over all years for one site.
-dDynamicOccupancy <- nimbleFunction(
-  ## I've checked that this runs and compiles, but I haven't tested if
-  ## I got the logic right!
-  run = function(x = double(2),
-    nrep = integer(),
-    psi1 = double(),
-    phi = double(1),
-    gamma = double(1),
-    p = double(1),
-    log = integer(0, default = 0)) {
-    prob1 <- psi1 * p[1]
-    numObs <- sum(x[,1]) ## do I have the right orientation?
-    ## prob of the occupied sites out of the total sites given p
-    ProbOccAndCount <- psi1 * dbinom(numObs, size = nrep, p = p[1], log = 0)
-    ## prob of the empty sites
-    ProbUnoccAndCount <- (1-psi1) * (numObs == 0)
-    ## probably of the observed states
-    ProbCount <- ProbOccAndCount + ProbUnoccAndCount
-    ProbOccGivenCount <- ProbOccAndCount / ProbCount
-    ProbOccNextTime <- ProbOccGivenCount * phi[1] +
-      (1-ProbOccGivenCount) * gamma[1]
-    ll <- log(ProbCount)
-    nyears <- dim(x)[2]
-    for(t in 2:nyears) {
-      numObs <- sum(x[,t])
-      ProbOccAndCount <- ProbOccNextTime *
-        dbinom(numObs, size = nrep, p = p[t], log = 0)
-      ProbUnoccAndCount <- (1-ProbOccNextTime) * (numObs == 0)
-      ProbCount <- ProbOccAndCount + ProbUnoccAndCount
-      ProbOccGivenCount <- ProbOccAndCount / ProbCount
-      ll <- ll + log(ProbCount)
-      if(t < nyears) ProbOccNextTime <- ProbOccGivenCount * phi[t] +
-        (1-ProbOccGivenCount) * gamma[t]
-    }
-    if(log) return(ll)
-    else return(exp(ll))
-    returnType(double())
-  }
-  )
-
-rDynamicOccupancy <- nimbleFunction(
[…]

### 2016-10-19 analysis/cppp/testcppp.R
attempted ot debug corss validation unction

@@ -1,7 +1,7 @@
 rm(list=ls())
 library(nimble)
 library(parallel)
-options(mc.cores=2)
+options(mc.cores=1)
 nthin <- 2
 
 pumpCode <- nimbleCode({

### 2016-10-19 analysis/crossValidation/crossValidationFunction.R

@@ -1,4 +1,5 @@
 library(nimble)
+library(coda)
 library(parallel)
 
 crossValCalculate <- function(row, MCMCOut, dataDimensions, saveData){
@@ -57,19 +58,29 @@ crossValidateOne <- function(model,
     C.modelMCMC$run(MCMCIter)
     MCMCout <- as.matrix(C.modelMCMC$mvSamples)
     sampNum <- dim(MCMCout)[1] 
-    crossValValue <- unlist(mclapply(ceiling(burnIn/thin):sampNum,
+    crossValValue <- sapply(ceiling(burnIn/thin):sampNum,
                                      crossValCalculate, MCMCout,
-                                     dataDimensions, saveData))
+                                     dataDimensions, saveData)
+    print(i)
     crossValAverage <- log(mean(crossValValue))
     nimble:::clearCompiled(C.modelMCMC)
-    return(crossValAverage)
+    return(list(crossValAverage =crossValAverage,
+                samples= MCMCout))
   }
-  crossVal <- sum(sapply(1:numBlocks, calcCrossVal,
+  crossValOut <- mclapply(1:numBlocks, calcCrossVal,
                            saveData,
                            newModel,
                            leaveOutIndex,
                            dataDimensions,
-                           dataNames))
+                           dataNames)
+  browser()
+  crossVal <- sum(sapply(crossValOut, function(x) x$crossValAverage))
+  samples <- lapply(crossValOut, function(x) x$samples)
+  chain.diag <-  do.call(rbind,
+                         lapply(lapply(samples, geweke.diag),
+                                function(x) x$z))
 
-  return(crossVal)
+  return(list(crossVal=crossVal,
+              chain.diag=chain.diag,
+              samples=samples))
 }

### 2016-10-19 analysis/crossValidation/crossValidationFunction_test.R

@@ -31,7 +31,16 @@ data <- cbind(rnorm(6, 0, 1), rnorm(6, 6, 1), rnorm(6, 4, 1),
 
 dyesModel$setData(list(y = data))
 
-output <- crossValidateOne(dyesModel, "y", 1000, 300, 2, 2)
+options(cores=1)
+niter <- 10
+output <- crossValidateOne(model=dyesModel,
+                           dataNames= "y",
+                           MCMCIter= niter,
+                           burnIn=niter*0.1,
+                           thin=1,
+                           leaveOutIndex=2,
+                           MCMCdefs=NULL)
+
 
 ## ****************************************************
 dyesCodeSimp <- nimbleCode({

### 2016-10-19 analysis/multiSpp-singleSea/opt1-4.R

@@ -110,7 +110,6 @@ ms.ss.occ <- nimbleCode({
   ## }
 })
 
-
 input1 <- c(code=ms.ss.occ, model.input)
 
 
@@ -288,7 +287,8 @@ occ.R.model <- nimbleModel(code=ms.ss.occ,
                            inits=input1$inits,
                            check=FALSE)
 
-options(mc.cores=1)
+niter <- 10
+options(mc.cores=2)
 
 source('../crossValidation/crossValidationFunction.R')
 

### 2016-10-19 analysis/spatial/src/dataGen.R

@@ -15,7 +15,7 @@ genSpatialOccData <- function(ngrid = 50,
                               beta1 = 6,
                               p = 0.8,
                               sigma = 0.5,
-                              delta = 0.5){
+                              delta = 1.5){
 
   ## Set up a square lattice region
   simgrid <- expand.grid(1:ngrid, 1:ngrid)

### 2016-10-19 ms/ms/ms.tex
misc

@@
   \begin{enumerate}
-  \item occupancy models are everywhere, but model fitting and
-    assessment are extremely computationally intensive
+  \item Something expressing the sentiment: occupancy models are
+    everywhere, but model fitting and assessment are extremely
+    computationally intensive
   \item Because models are so computationally intensive, users often
@@
 Estimating the proportion of sites occupied by a species is common
-challenge for many sub-disciplines ecology and evolution including
+challenge for many sub-disciplines in ecology and evolution including
 meta-population, endangered species and invasion biology. Greater
@@
 occupancy or shared variation in occupancy across species is often of
-greatest interest to ecologists \citep[e.g.,][]{iknayan2014detecting},
+greatest interest to biologists \citep[e.g.,][]{iknayan2014detecting},
 the Hidden Markov Model is often embedded within a hierarchical
@@
 Monte Carlo (MCMC) to perform a Bayesian analysis. Standard MCMC
-software will including the latent state variables in MCMC sampling
+software will include the latent state variables in MCMC sampling
 \citep[e.g.,][]{plummer-2003-jags, winbugs, openbugs}. Such models are
@@
 In addition, fitting these models is such a challenge that users often
-forgo adding any additional computation to asses model fit. A common
-idea behind evaluating whether a model provides an adequate fit to a
+forgo any additional computation to asses model fit. A common idea
+behind evaluating whether a model provides an adequate fit to a
 dataset is that if data is simulated from the model, the simulated
@@
 the most widely used applications of statistics by
-practitioners. Though many methods for Bayesian model selection such
-as cross-validation have been developed \citep{hooten2014guide}, they,
-like model assessment, are computationally intensive --- particularly
-for hierarchical models like occupancy models. A typical need for
-model selection arises when a practitioner is choosing whether to
+practitioners. Though many methods for Bayesian model selection have
+been developed \citep{hooten2014guide}, but they, like model
+assessment, are computationally intensive. For example,
+cross-validation, one of the most fudamental procedures in model
+selection, requires iternativly re-fitting the model. A typical need
+for model selection arises when a practitioner is choosing whether to
 include a specific layer of hierarchy (i.e., random effect). This is
@@
 \citep{celeux2006deviance, hooten2014guide}, because it is built into
-open-source software such as WinBUGS \citep{winbugs}, it is
-uncritically used by practitioners. Readily available and
-theoretically sound alternative methods are thus critically needed.
+open-source software such as WinBUGS, it can be used uncritically by
+practitioners. Readily available and theoretically sound alternative
+methods are thus greatly needed.
 
-Luckily, methods to improve MCMC efficiency of Hidden Markov Models,
-such filtering over latent states to calculate model likelihoods
-in order to limit MCMC sampling to top-level parameters dynamic
-blocking or parameters \citep{turek2016efficient}, have been
-developed. A synergistic strategy is to assign specific MCMC samplers
-to different parameters depending on the nature of those nodes (i.e.,
-discrete versus continuous). Though these methods are available in
-isolation in application-specific software, they cannot be used in
-combination for any arbitrary model structure.
+Luckily, methods to improve MCMC efficiency of Hidden Markov Models
+have been developed such filtering over latent states to calculate
+model likelihoods in order to limit MCMC sampling to top-level
+parameters dynamic blocking or parameters
+\citep{turek2016efficient}. A synergistic strategy is to assign
+specific MCMC samplers to different parameters depending on the nature
+of those nodes (i.e., discrete versus continuous). Though these
+methods are available in isolation in application-specific software,
+they cannot be used in combination for any arbitrary model structure.
 
@@
 $z_{i}$ denote the true occupancy of a species at site $i$.  We then
-let $x_{i,j}$ indicate whether we detected ($x_{i,j}=1$) or did not
-detect ($x_{i,j}=0$) that species in the $j^{\mathrm{th}}$ visit to
-site $i$.  We assumed that occupancy at the $i^{\mathrm{th}}$ site is
-a Bernoulli random variable $z_{i} \sim \mathrm{Bern}(\psi_{i})$ with
-probability $\psi_{i}$.  We included the effect of an arbitrary
-covariate (e.g., elevation) on site occupancy. To model the spatial
-auto-correlation in occupancy between sites, we assume the co-variance
-between sites $Y_i$ and $Y_j$ is a function of distance between $p_i$
-and $p_j$. We computed the probability of occupancy at site $i$
+let $x_{i,j}$ indicate whether the species was ($x_{i,j}=1$), or was
+not detected ($x_{i,j}=0$) in the $j^{\mathrm{th}}$ visit to site $i$.
+We assumed that occupancy at the $i^{\mathrm{th}}$ site is a Bernoulli
+random variable $z_{i} \sim \mathrm{Bern}(\psi_{i})$ with probability
+$\psi_{i}$.  We included the effect of an arbitrary covariate (e.g.,
+elevation) on site occupancy. To model the spatial auto-correlation in
+occupancy between sites, we assume the co-variance between sites $Y_i$
+and $Y_j$ is a function of distance between $p_i$ and $p_j$. We
+computed the probability of occupancy at site $i$
 %
@@
   model...}
+
+We implemented a procedure to calculate calibrated posterior
+predictive $p$-values \citep[CPPP, ][]{hjort-etal-06} to assesss the
+fit of the model to the data. After the parameters have been fit to
+the model, a sample of the posterior is used to simulate data from the
+model. A discrepancy measure, which we chose to be the model
+likelihood, is then calculated, and the posterior $p$-value is the
+number of simulated $p$-values that fall below the observed. To
+``calibrate'' the distribution of posterior $p$-values, the MCMC is
+rerun on the simulated data to refit the model. If the CPPP $< 0.05$,
+the model is rejected as having an adequate fit ot the data
+\citep{hjort-etal-06}.
 
@@
 
-We then let $x_{i,j,k}$ indicate whether we detected ($x_{i,j,k}=1$)
-or did not detect ($x_{i,j,k}=0$) that species in the
+We then let $x_{i,j,k}$ indicate whether that species was
+($x_{i,j,k}=1$) or was not detected ($x_{i,j,k}=0$) in the
 $k^{\mathrm{th}}$ visit to site $i$ in year $j$. We assume detection
@@
 (``NIMBLE-latent'' and ``JAGS-latent'', respectively). Following
-\cite{Royle2007, kery-schaub-11}, we use uninformative
+\cite{Royle2007, kery-schaub-11}, we use uninformative,
 $\mathrm{Unif}(0,1)$ priors for all parameters.
@@
 
+As with the spatail model, we use calibrated posterior predictive
+$p$-values \citep{hjort-etal-06} to assesss the fit of the model to
+the data.
+
 \subsubsection*{Multi species, single season occupancy model}
@@
 common distribution with an estimated variance.  For species $i$, we
-let $z_{i,j}$ denote its true occupancy state at site $j$.  We then
-let $x_{i,j,k}$ indicate whether we detected ($x_{i,j,k}=1$) or did
-not detect ($x_{i,j,k}=0$) that species in the $k^{\mathrm{th}}$ visit
-to site $j$.  We assumed that the occupancy of the $i^{\mathrm{th}}$
-species at the $j^{\mathrm{th}}$ site is a Bernoulli random variable
-$z_{i,j} \sim \mathrm{Bern}(\psi_{i,j})$. We then let $x_{i,j,k}$
-indicate whether we detected ($x_{i,j,k}=1$) or did not detect
-($x_{i,j,k}=0$) species $i$ in the $k^{\mathrm{th}}$ visit to site
-$j$. We also assumed that detection was distributed according to be a
-Bernoulli random variable such that $x_{i,j,k} \sim
-\mathrm{Bern}(p_{i}*z_{i,j})$, where $p_{i}$ is the probability that
-the $i^{\mathrm{th}}$ species was detected. Both site occupancy and
-detection were influence by habiat and survey characteristics
-\citep{zipkin2010multi}. Specifically, occurence depended on the study
-area (CATO, Ind=1, or FCW, Ind=0), the basal tree area (BA) and the
-understory foliage cover (UFC). The species-specific occupancy
-probabilities are modeled as
+let $z_{i,j}$ denote its true occupancy state at site $j$. We assumed
+that the occupancy of the $i^{\mathrm{th}}$ species at the
+$j^{\mathrm{th}}$ site is a Bernoulli random variable $z_{i,j} \sim
+\mathrm{Bern}(\psi_{i,j})$.  We then let $x_{i,j,k}$ indicate whether
+species $i$ was ($x_{i,j,k}=1$) or was not detected ($x_{i,j,k}=0$) in
+the $k^{\mathrm{th}}$ visit to site $j$. We also assumed that
+detection was distributed according to be a Bernoulli random variable
+such that $x_{i,j,k} \sim \mathrm{Bern}(p_{i}*z_{i,j})$, where $p_{i}$
+is the probability that the $i^{\mathrm{th}}$ species was
+detected. Both site occupancy and detection were influence by habiat
+and survey characteristics \citep{zipkin2010multi}. Specifically,
+occurence depended on the study area (CATO, Ind=1, or FCW, Ind=0), the
+basal tree area (BA) and the understory foliage cover (UFC). The
+species-specific occupancy probabilities are modeled as
 
@@
     \alpha1_iUFC_j + \alpha2_iUFC^2_j + \alpha3_iBA_j +
-    \alpha4_iBA^2_j \hspace{0.2em}
+    \alpha4_iBA^2_j \\
+     \alpha1 &\sim N({\mu}_{\alpha1}, \sigma^2_{\alpha1})\\
+     \alpha2 &\sim N({\mu}_{\alpha2}, \sigma^2_{\alpha2})\\
+     \alpha3 &\sim N({\mu}_{\alpha3}, \sigma^2_{\alpha3})\\
+     \alpha4 &\sim N({\mu}_{\alpha4}, \sigma^2_{\alpha4})
+   \hspace{0.2em}
   \end{split}
@@
 
-Similarly, detection survey location as well as the date and time
-since sunrise.
+Similarly, detection depended on survey location and the date:
 
@@
     \mathrm{logit}(p_{i,j,k}) &= vCATO_i(Ind_j) + vFCW_i(1-Ind_j) +
-    \beta1_idate_j + \beta2_idate^2_j + \beta3_isunrise_j
+    \beta1_idate_j + \beta2_idate^2_j\\
+     \beta1 &\sim N({\mu}_{\beta1}, \sigma^2_{\beta1})\\
+     \beta2 &\sim N({\mu}_{\beta2}, \sigma^2_{\beta2})\\
     \hspace{0.2em}
@@
 (``NIMBLE-latent'' and ``JAGS-latent'', respectively). We use
-uninformative priors $\mathrm{Norm}(0,1000)$ for the means of the
+uninformative priors, $\mathrm{Norm}(0,1000)$ for the means of the
 distributions of the hyperparameters and $\mathrm{Unif}(0,100)$ the
@@
 states to calculate model likelihoods in order to limit MCMC sampling
-to top-level parameters (``Filter''). We also vectorized all
+to top-level parameters (``filter''). We also vectorized all
 calculations that would have require for loops in JAGS. \flagged{Do we
@@
 to speed sampling of the top-level parameters 1) dynamic blocking of
-the parameters \citep[``Filter + autoblocking'',
+the parameters \citep[``filter + autoblocking'',
 ][]{turek2016efficient}, and 2) a custom blocking scheme where the
@@
 
-\subsection*{Model assessment}
-\label{sec:assess}
-
-We implemented a procedure to calculate calibrated posterior
-predictive $p$-values \citep{hjort-etal-06}. After the parameters have
-been fit to the model, a sample of the posterior is used to simulate
-data from the model. A discrepancy measure, which we chose to be the
-model likelihood, is then calculated, and the posterior $p$-value is
-the number of simulated $p$-values that fall below the observed. To
-``calibrate'' the distribution of posterior $p$-values, the MCMC is
-rerun on the simulated data to refit the model.
-
-\subsection*{Model selection}
-\label{sec:select}
-
-Cross-validation is one of the most fudamental procedures in model
-selection, but, because it requires iternativly re-fitting the model,
-is computationally intensive \citep{hooten2014guide}. In cross
-validation, we exclude a subset of the data ($y_k$) from model
-fitting, then use the fitted model to predict $y_k$. The prediction
-error is summarized by comparing the simulated $y_k$ to the true
-$y_k$.
-
-In multi-species occupancy models like the model we explored in
-Section \ref{sec:msss}), practitioners are often interested in
+We also use calibrated posterior predictive $p$-values
+\citep{hjort-etal-06} to assesss the fit of the model to the data.  In
+multi-species occupancy models, practitioners are often interested in
 determining whether a model including a species random effect for
@@
 were included.
+
+% a subset of the data ($y_k$) if excluded from model fitting, the the
+% fitted model is used to predict $y_k$. The prediction error is
+% summarized by comparing the simulated $y_k$ to the true $y_k$.
 

### 2016-10-19 analysis/all/plotting.R

@@ -88,11 +88,46 @@ plotEffSize <- function(eff.size, eff.param,
     mtext("Effective sample size \n per second (log)",
           2, line=3.2, cex=1.5)
   }
+
+    f3 <- function(){
+    layout(matrix(1:2, ncol=1), heights=c(1,3))
+
+    par(oma=c(5, 6, 0.5, 1),
+        mar=c(0.5, 0, 0.5, 1), cex.axis=1.5)
+    
+    diffs <- t(apply(eff.param, 1,
+                   function(x) x - eff.param["JAGS-latent",]))[-1,]
+    
+    plot(NA, ylim=c(0,1), xlim=c(0,1), yaxt= "n", xaxt="n", bty="n")
+    legend("top", legend=rownames(diffs),
+           pch=16, col=cols[-1], bty="n", ncol=3)
+    
+    plot(NA, ylim=range(diffs), xlim=c(1, ncol(diffs)),
+         xlab="", ylab="", xaxt="n")
+    for(i in 1:nrow(diffs)){
+      points(x=1:ncol(diffs), y=diffs[i,],
+             col=cols[i+1],
+             pch=16,
+             xaxt="n",
+             type="o")
+    }
+    text(x=1:ncol(diffs), par('usr')[3],
+         srt = 45, adj = 1 + adj2,
+         labels = colnames(diffs),
+         xpd = NA,
+         cex=0.8)
+    mtext("Difference in effective sample size \n per second",
+          2, line=3.2, cex=1.5)
+    abline(h=0, lty=2)
+  }
   pdf.f(f,
         file= file.path(sprintf(f.path, name, "Bar")),
         height=6, width=widths[1])
   pdf.f(f2,
         file= file.path(sprintf(f.path, name, "Points")),
+        height=4, width=widths[2])
+  pdf.f(f3,
+        file= file.path(sprintf(f.path, name, "Diffs")),
         height=4, width=widths[2])
 }
 

### 2016-10-19 analysis/crossValidation/crossValidationFunction.R

@@ -74,7 +74,8 @@ crossValidateOne <- function(model,
                            dataDimensions,
                            dataNames)
   browser()
-  crossVal <- sum(sapply(crossValOut, function(x) x$crossValAverage))
+  crossVal <- sum(sapply(crossValOut, function(x) x$crossValAverage),
+                  na.rm=TRUE)
   samples <- lapply(crossValOut, function(x) x$samples)
   chain.diag <-  do.call(rbind,
                          lapply(lapply(samples, geweke.diag),

### 2016-10-19 analysis/crossValidation/crossValidationFunction_test.R

@@ -31,8 +31,8 @@ data <- cbind(rnorm(6, 0, 1), rnorm(6, 6, 1), rnorm(6, 4, 1),
 
 dyesModel$setData(list(y = data))
 
-options(cores=1)
-niter <- 10
+options(mc.cores=1)
+niter <- 100
 output <- crossValidateOne(model=dyesModel,
                            dataNames= "y",
                            MCMCIter= niter,

### 2016-10-19 analysis/multiSpp-singleSea/compare.R

@@ -1,5 +1,4 @@
 rm(list=ls())
-
 library(nimble)
 library(coda)
 
@@ -74,7 +73,6 @@ make_MCMC_comparison_pages(ms.ss.occ.all,
 by.param <- apply(ms.ss.occ.all[[1]]$samples, c(1,2), effectiveSize)/
   ms.ss.occ.all[[1]]$timing 
 by.config <- ms.ss.occ.all[[1]]$efficiency
-
 
 source('../all/plotting.R')
 plotEffSize(by.config, by.param, f.path= file.path(save.dir,

### 2016-10-19 analysis/multiSpp-singleSea/opt1-4.R

@@ -19,19 +19,21 @@ ms.ss.occ <- nimbleCode({
   ## Define prior distributions for community-level model parameters
   cato.occ.mean ~ dunif(0,1)
   mu.ucato <- log(cato.occ.mean) - log(1-cato.occ.mean)
-
+  sigma.ucato ~ dunif(0, 100)
+  
   fcw.occ.mean ~ dunif(0,1)
   mu.ufcw <- log(fcw.occ.mean) - log(1-fcw.occ.mean)
+  sigma.ufcw ~ dunif(0, 100)
 
   cato.det.mean ~ dunif(0,1)
   mu.vcato <- log(cato.det.mean) - log(1-cato.det.mean)
+  sigma.vcato ~ dunif(0, 100)
 
   fcw.det.mean ~ dunif(0,1)
   mu.vfcw <- log(fcw.det.mean) - log(1-fcw.det.mean)
+  sigma.vfcw ~ dunif(0, 100)
 
   ## random effects
-  sigma.ucato ~ dunif(0, 100)
-  sigma.ufcw ~ dunif(0, 100)
   
   mu.a1 ~ dnorm(0, 0.001)
   sigma.a1 ~ dunif(0, 100)
@@ -45,11 +47,9 @@ ms.ss.occ <- nimbleCode({
   mu.a4 ~ dnorm(0, 0.001)
   sigma.a4 ~ dunif(0, 100)
 
-  sigma.vcato ~ dunif(0, 100)
-  sigma.vfcw ~ dunif(0, 100)
-  
   mu.b1 ~ dnorm(0, 0.001)
   sigma.b1 ~ dunif(0, 100)
+  
   mu.b2 ~ dnorm(0, 0.001)
   sigma.b2 ~ dunif(0, 100)
 
@@ -360,10 +360,10 @@ ms.ss.occ.simp <- nimbleCode({
 
 
 occ.R.model.simp <- nimbleModel(code=ms.ss.occ.simp,
-                           constants=input1$constants,
-                           data=input1$data,
-                           inits=input1$inits,
-                           check=FALSE)
+                                constants=input1$constants,
+                                data=input1$data,
+                                inits=input1$inits,
+                                check=FALSE)
 
 occ.mcmc <- buildMCMC(occ.R.model.simp)
 occ.C.model.simp <- compileNimble(occ.R.model.simp)
@@ -371,9 +371,9 @@ occ.C.mcmc <- compileNimble(occ.mcmc, project = occ.R.model.simp)
 
 
[…]

### 2016-10-19 analysis/spatial/src/dataGen.R

@@ -9,13 +9,13 @@ rmvn <- function(n, mu = 0, V = matrix(1)) {
 }
 
 
-genSpatialOccData <- function(ngrid = 50,
+genSpatialOccData <- function(ngrid = 80,
                               nreps = 50,
                               alpha = 1,
                               beta1 = 6,
                               p = 0.8,
-                              sigma = 0.5,
-                              delta = 1.5){
+                              sigma = 2,
+                              delta = 0.5){
 
   ## Set up a square lattice region
   simgrid <- expand.grid(1:ngrid, 1:ngrid)
@@ -27,7 +27,7 @@ genSpatialOccData <- function(ngrid = 50,
   ## Generate spatial random effect
   X <- rmvn(1, rep(0, n),  (sigma^2)*exp(-delta * distance))
   Xraster <- rasterFromXYZ(cbind(simgrid[, 1:2] - 0.5, X))
-
+  plot(Xraster)
   ## simulate elevation data
   elev <- raster(matrix(rnorm(n), ngrid, ngrid),
                  xmn = 0, xmx = ngrid,

### 2016-10-31 ms/ms/ms.tex
added figures to ms

@@
 \usepackage{graphicx}
+\usepackage{rotating}
 \usepackage{textcomp}
-\usepackage{xr}
-
 \hypersetup{colorlinks=true, linkcolor=black, citecolor=black}
@@
 
-\def\title{Something occupancy models}
-\def\author{Lauren C.\ Ponisio$^{1,2}$, Nicholas Michaud$^1$, Perry de
+\def\title{Improving the efficiency of occupancy models (something
+  more interesting sounding?)}
+\def\author{Lauren C.\ Ponisio$^{1,2,3}$, Nicholas Michaud$^1$, Perry de
   Valpine$^1$ (Daniel and Chris?)}
@@
     94720\\
+  \item Berkeley Institute for Data Science (BIDS)\\
+    University of California, Berkeley\\
+    190 Doe Library \\
   \item Department of Entomology\\
@@
     sampling efficiency over, particularly with models including
-    random effects. (more results once they are available)
+    random effects (maybe?). (more results once they are available)
   \item \textit{Implications:} Ours results highlight the need for
@@
 assessment, are computationally intensive. For example,
-cross-validation, one of the most fudamental procedures in model
-selection, requires iternativly re-fitting the model. A typical need
+cross-validation, one of the most fundamental procedures in model
+selection, requires iterativly re-fitting the model. A typical need
 for model selection arises when a practitioner is choosing whether to
@@
 We implemented a procedure to calculate calibrated posterior
-predictive $p$-values \citep[CPPP, ][]{hjort-etal-06} to assesss the
+predictive $p$-values \citep[CPPP, ][]{hjort-etal-06} to assess the
 fit of the model to the data. After the parameters have been fit to
@@
 rerun on the simulated data to refit the model. If the CPPP $< 0.05$,
-the model is rejected as having an adequate fit ot the data
+the model is rejected as having an adequate fit to the data
 \citep{hjort-etal-06}.
@@
 
-As with the spatail model, we use calibrated posterior predictive
-$p$-values \citep{hjort-etal-06} to assesss the fit of the model to
+As with the spatial model, we use calibrated posterior predictive
+$p$-values \citep{hjort-etal-06} to assess the fit of the model to
 the data.
@@
 is the probability that the $i^{\mathrm{th}}$ species was
-detected. Both site occupancy and detection were influence by habiat
+detected. Both site occupancy and detection were influence by habitat
 and survey characteristics \citep{zipkin2010multi}. Specifically,
-occurence depended on the study area (CATO, Ind=1, or FCW, Ind=0), the
+occurrence depended on the study area (CATO, Ind=1, or FCW, Ind=0), the
 basal tree area (BA) and the understory foliage cover (UFC). The
@@
 We also use calibrated posterior predictive $p$-values
-\citep{hjort-etal-06} to assesss the fit of the model to the data.  In
+\citep{hjort-etal-06} to assess the fit of the model to the data.  In
 multi-species occupancy models, practitioners are often interested in
@@
 where the detection data for species is left out, the model refitted,
-and the fitted model used to predict the occurence of that
+and the fitted model used to predict the occurrence of that
 species. The predictive error of the model included a random effect of
@@
 \label{sec:results}
+
+\subsubsection*{Multi species, single season occupancy model}
+\label{sec:msss-res}
+
 \subsubsection*{Single species, single season occupancy model with
   spatial auto-correlation}
-
-\subsubsection*{Multi species, single season occupancy model}
-\label{sec:msss}
+\label{sec:sp-res}
 
@@
   model}
+\label{sec:ssms-res}
 
@@
 
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=.8\textwidth]{../analysis/changePoint/plotting/networks.pdf}
-%   \caption{Assembling hedgerow networks had more changing points
-%     (vertical red lines) than non-assembling hedgerows and weedy field
-%     margins (a representative sample of non-assembling sites are
-%     depicted here). In each network, plants and pollinators are
-%     represented by green and yellow circles, respectively, weighted by
-%     their degree. Each species has a consistent position in the
-%     network across years. In the assembling hedgerows, colored squares
-%     in the corner of each network represent the year of
-%     assembly. Before hedgerows are installed (when they are still
-%     field margins) the year of assembly is zero.}
-%   \label{fig:changePoints}
-% \end{figure}
-% \clearpage
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\textwidth]{../../../saved/SingleSpp-MultiSea/figures/comparisons/SingleSpp-MultiSeaBar.pdf}
+  \caption{The mean and minimum efficiency of different MCMC samplers
+    for the single species, multi season occupancy model.}
+  \label{fig:ss-ms-eff}
+\end{figure}
+\clearpage
+
+\begin{sidewaysfigure}
+  \centering
+  \includegraphics[width=.8\textwidth]{../../../saved/SingleSpp-MultiSea/figures/comparisons/SingleSpp-MultiSeaDiffs.pdf}
+  \caption{The log ratio of the NIMBLE samplers in comparison to the
+    JAGS sampler for the single species, multi season occupancy
+    model.}
+  \label{fig:ss-ms-eff}
+\end{sidewaysfigure}
+\clearpage
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\textwidth]{../../../saved/multiSpp-singleSea/figures/comparisons/MultiSpp-SingleSeaBar.pdf}
+  \caption{The mean and minimum efficiency of different MCMC samplers
+    for the multi species, single season occupancy model.}
+  \label{fig:ms-ss-eff}
+\end{figure}
+\clearpage
+
+\begin{sidewaysfigure}
+  \centering
+  \includegraphics[width=.8\textwidth]{../../../saved/multiSpp-singleSea/figures/comparisons/MultiSpp-SingleSeaDiffs.pdf}
+  \caption{The log ratio of the NIMBLE samplers in comparison to the
+    JAGS sampler for the multi species, single season occupancy
+    model.}
+  \label{fig:ms-ss-eff}
+\end{sidewaysfigure}
+\clearpage
 

### 2016-10-31 analysis/all/plotting.R

@@ -96,7 +96,7 @@ plotEffSize <- function(eff.size, eff.param,
         mar=c(0.5, 0, 0.5, 1), cex.axis=1.5)
     
     diffs <- t(apply(eff.param, 1,
-                   function(x) x - eff.param["JAGS-latent",]))[-1,]
+                   function(x) log(x) - log(eff.param["JAGS-latent",])))[-1,]
     
     plot(NA, ylim=c(0,1), xlim=c(0,1), yaxt= "n", xaxt="n", bty="n")
     legend("top", legend=rownames(diffs),
@@ -116,7 +116,7 @@ plotEffSize <- function(eff.size, eff.param,
          labels = colnames(diffs),
          xpd = NA,
          cex=0.8)
-    mtext("Difference in effective sample size \n per second",
+    mtext("Ratio of effective sample size \n per second (log)",
           2, line=3.2, cex=1.5)
     abline(h=0, lty=2)
   }

### 2016-10-31 analysis/singleSpp-multiSea/main.R

@@ -4,6 +4,7 @@ setwd("~//Dropbox//nimble//occupancy//analysis//singleSpp-multiSea")
 ## original model
 source('original.R')
 
+
 ## custom samplers for zs and reflective sampler, or slice sampler
 source('opt2-3.R')
 

### 2016-10-31 analysis/singleSpp-multiSea/src/dataGen.R

@@ -21,7 +21,7 @@ expit <- function(x) {
 ## range.p - bounds of uniform distribution from which annual p drawn
 ## range.psi and range.gamma - same for survival and colonization probability
 
-genDynamicOccData <- function(R = 250,
+genDynamicOccData <- function(R = 50,
                               J = 10, K = 10,
                               psi1 = 0.4,
                               range.p = c(0.2, 0.4),

### 2018-03-11 ms/notes/jags2nimble/jags2nimble.tex
added multi season mutli species model

\documentclass[12pt]{article}
\usepackage[top=0.6in, bottom=0.6in, right=0.7in, left=0.7in,
paperwidth=8.5in, paperheight=11in, nohead]{geometry}
\geometry{letterpaper}
\usepackage[pdftex]{graphicx}
\usepackage{color}
\usepackage[normalem]{ulem}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{setspace}
\usepackage{mdwlist}
\usepackage{verbatim}
\usepackage{tikz}
\usepackage{rotating}
\usepackage[numbers]{natbib}

%% make bibliography more compact \setlength{\bibsep}{0in}
\renewcommand\bibsection{\subsubsection*{\refname}}
\title{Transitioning from JAGS/BUGS to NIMBLE}
\author{L.C.~Ponisio, Perry de Valpine, Nicholas L. Michaud}
\def\code#1{\texttt{#1}}

\begin{document}

\maketitle

\section{Steps}
\begin{enumerate}
\item Wrap your
  JAGS/BUGS model code in \code{nimbleCode(\{\})}, directly in R. This
  replaces the step of generating a file containing the model code, as
  is often done
  via \code{cat} or \code{sink} prior to calling BUGS/JAGS
  via \code{R2WinBUGS}, \code{R2jags},
  \code{rjags} or \code{jagsUI}.  (Alternatively, one can read
  JAGS- and BUGS-formatted code and data files using
  \code{readBUGSmodel}.)
\item Provide information about empty indices.  The easiest solution
  is to write the index range explicitly. For example, \code{Omega[,]} will not work in NIMBLE, but
  \code{Omega[1:nsite,1:nspecies]} will.  (Alternatively, one can use
  the \code{dimensions} argument to \code{nimbleModel}.)
\item Split the \code{data} for BUGS/JAGS into \code{data} and
  \code{constants} for NIMBLE.  Constants are necessary to define the
  model, such as \code{nsite} in \code{for(i in 1:nsite) \{...\}}.  Data are
  observed values of some variables, which in NIMBLE can be changed
  and do not need to be provided when a model is created via
  \code{nimbleModel}.  (Alternatively, one can provide a list of both
  constants and data for the \code{constants} argument to
  \code{nimbleModel}, and NIMBLE will determine which is which.)
\item Convert \code{inits} and \code{monitors} inputs from functions to named lists.
\item To run an MCMC on your model, you have two choices:
  \begin{itemize}
  \item Use \code{nimbleMCMC()} much like a call to
  \code{jags()} from the R2jags package.  This will take all steps to
  set up and run an MCMC using
   NIMBLE's default configuration (Figure \ref{fig:wf}). 
 \item To use NIMBLE's full flexibiilty, build the model, configure
   and build the MCMC, and compile both the model and MCMC before
   running the MCMC. Each of
   these pieces can be customized
  and/or explored individually (Figure \ref{fig:wf}).
  \end{itemize}
\end{enumerate}

\section{Differences}
\begin{enumerate}
\item NIMBLE models allow alternative parameterizations for
  distributions (e.g. \code{sd} instead of \code{tau}), vectorized
  math, and user-defined functions and distributions.  
\item NIMBLE allows an MCMC configuration to be customized by changing
  the the samplers to be used.
\item NIMBLE allows new samplers and other kinds of algorithms to be
  programmed from R.
\item \code{burnin} in NIMBLE is post-thinning -- whereas it is pre-thinning in
  BUGS/JAGS -- when using the \code{runMCMC} function.
\end{enumerate}

\clearpage
\begin{sidewaysfigure}
  \centering
  \includegraphics[width=.8\textwidth]{figures/nimbleWorkflow.pdf}
  \caption{Workflow options within NIMBLE.}
  \label{fig:wf}
\end{sidewaysfigure}
\clearpage

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-PDF-mode: t
%%% End:

### 2018-03-11 analysis/all/plotting.R

@@ -10,7 +10,7 @@ pdf.f <- function(f, file, ...) {
 
 checkChains <- function(all.mods.samps, f.path,
                         only.one="jags",
-                        params=NULL){
+                        params=NULL, prop.plot=0.1){
     ## function to plot values of parameters as a function of
     ## iterations. Input is a MCMCcompare object
     niter <- dim(all.mods.samps)[3]
@@ -24,7 +24,7 @@ checkChains <- function(all.mods.samps, f.path,
         f <- function(){
             layout(matrix(1:4, ncol=2))
             apply(all.mods.samps[z,params,], 1, function(x){
-                plot(x[seq(from=1, to=length(x), length.out=1000)], type="l",
+                plot(x[seq(from=1, to=length(x), length.out=length(x)*prop.plot)], type="l",
                      xlab = 'iteration',
                      main= params[which(apply(all.mods.samps[z,,], 1,
                                               function(y)
@@ -51,7 +51,7 @@ plotEffSize <- function(eff.size, eff.param,
             mar=c(0.5, 0, 2.5, 1), cex.axis=1.5)
         ## barplots
         mp1 <- barplot(eff.size$mean, names="", las=1, col=cols)
-        mtext("Mean", 3, line=0.5, cex=1.5)
+        mtext("Mean", 3, line=0.5, cex=1.2)
 
         mp2 <- barplot(eff.size$min, names="", las=1, col=cols)
         text(mp2, par('usr')[3] - adj1,
@@ -60,7 +60,7 @@ plotEffSize <- function(eff.size, eff.param,
              xpd = NA,
              cex=1)
 
-        mtext("Minimum", 3, line=0.5, cex=1.5)
+        mtext("Minimum", 3, line=0.5, cex=1.2)
         mtext("Effective sample size \n per second",
               2, line=4.5, cex=1.5, at=at)
     }
@@ -73,7 +73,7 @@ plotEffSize <- function(eff.size, eff.param,
 
         plot(NA, ylim=c(0,1), xlim=c(0,1), yaxt= "n", xaxt="n", bty="n")
         legend("top", legend=rownames(eff.param),
-               pch=16, col=cols, bty="n", ncol=3)
+               pch=16, col=cols, bty="n", ncol=3, cex=0.7)
 
         plot(NA, ylim=log(range(eff.param)), xlim=c(1, ncol(eff.param)),
[…]

### 2018-03-11 analysis/multiSpp-multiSea/filtering.R

@@ -1,0 +1,26 @@
+## ************************************************************
+## setwd('~/Dropbox/occupancy')
+rm(list=ls())
+setwd('analysis/multiSpp-multiSea')
+
+source("src/initialize.R")
+load("data/5-0-350.Rdata")
+source("src/complete_allInt_filtering.R")
+
+input1 <- c(code=ms.ms.occ,
+            model.input)
+
+## *****************************************************************
+## vanilla nimble/jags
+## *****************************************************************
+
+ms.ms.nimble <- compareMCMCs_withMonitors(input1,
+                                          MCMCs=c('nimble'),
+                                          niter=niter,
+                                          burnin = burnin,
+                                          summary=FALSE,
+                                          check=FALSE,
+                                          monitors=model.input$monitors)
+
+save(ms.ms.orig, file=file.path(save.dir, 'filtering.Rdata'))
+

### 2018-03-11 analysis/multiSpp-multiSea/original.R

@@ -1,0 +1,56 @@
+## ************************************************************
+## setwd('~/Dropbox/occupancy')
+rm(list=ls())
+setwd('analysis/multiSpp-multiSea')
+
+source("src/initialize.R")
+load("data/5-0-350.Rdata")
+source("src/complete_allInt.R")
+
+input1 <- c(code=ms.ms.occ,
+            model.input)
+
+## *****************************************************************
+## vanilla nimble/jags
+## *****************************************************************
+
+ms.ms.nimble <- compareMCMCs_withMonitors(input1,
+                                          MCMCs=c('nimble', 'jags'),
+                                          niter=niter,
+                                          burnin = burnin,
+                                          summary=FALSE,
+                                          check=FALSE,
+                                          monitors=model.input$monitors)
+
+save(ms.ms.orig, file=file.path(save.dir, 'orig.Rdata'))
+
+## *********************************************************************
+## sampler only a subset of latent states
+## *********************************************************************
+
+MCMCdefs.subsamp <- list('nimbleSubsamp' = quote({
+    customSpec <- configureMCMC(Rmodel)
+    customSpec$removeSamplers('Z')
+    customSpec$addSampler('Z', type = 'sampler_latentSub',
+                          control = list(leaveOutProportion = 0.05,
+                                         control = list()))
+    customSpec
+}))
+
+ms.ms.subsamp <- compareMCMCs_withMonitors(input1,
+                              MCMCs=c('nimbleSubsamp'),
+                              MCMCdefs = MCMCdefs.subsamp,
+                              niter= niter,
+                              burnin = burnin,
+                              monitors=model.input$monitors,
+                              summary=FALSE,
+                              check=FALSE)
+
+save(ms.ms.subsamp, file=file.path(save.dir, 'subsamp.Rdata'))
+
+
+## *********************************************************************
+## cross level sampler
+## *********************************************************************
+## sample latent and top-level parameters jointly
+

### 2018-03-11 analysis/multiSpp-multiSea/src/complete_allInt.R

@@ -1,0 +1,293 @@
+ms.ms.occ <- nimbleCode({
+    ## multi-species priors
+    ## detectablility
+    mu.p.0     ~ dnorm(0,0.01)
+    mu.p.day.1 ~ dnorm(0,0.01)
+    mu.p.day.2 ~ dnorm(0,0.01)
+
+    sigma.p.0     ~ dunif(0,10)
+    sigma.p.day.1 ~ dunif(0,10)
+    sigma.p.day.2 ~ dunif(0,10)
+
+    tau.p.0 <- 1/(sigma.p.0*sigma.p.0)
+    tau.p.day.1 <- 1/(sigma.p.day.1*sigma.p.day.1)
+    tau.p.day.2 <- 1/(sigma.p.day.2*sigma.p.day.2)
+
+    ## phi/gam random intercepts
+    mu.phi.0  ~ dnorm(0,0.01)
+    mu.gam.0  ~ dnorm(0,0.01)
+
+    sigma.phi.0 ~ dunif(0,10)
+    sigma.gam.0 ~ dunif(0,10)
+    tau.phi.0 <-  1/(sigma.phi.0*sigma.phi.0)
+    tau.gam.0 <-  1/(sigma.gam.0*sigma.gam.0)
+
+    ## hedgerow area proximity
+    mu.phi.hr.area  ~ dnorm(0,0.01)
+    mu.gam.hr.area  ~ dnorm(0,0.01)
+
+    sigma.phi.hr.area ~ dunif(0,10)
+    sigma.gam.hr.area ~ dunif(0,10)
+    tau.phi.hr.area <- 1/(sigma.phi.hr.area*sigma.phi.hr.area)
+    tau.gam.hr.area <- 1/(sigma.gam.hr.area*sigma.gam.hr.area)
+
+    ## semi nat habitat area proximity
+    mu.phi.nat.area  ~ dnorm(0,0.01)
+    mu.gam.nat.area  ~ dnorm(0,0.01)
+
+    sigma.phi.nat.area ~ dunif(0,10)
+    sigma.gam.nat.area ~ dunif(0,10)
+    tau.phi.nat.area <- 1/(sigma.phi.nat.area*sigma.phi.nat.area)
+    tau.gam.nat.area <- 1/(sigma.gam.nat.area*sigma.gam.nat.area)
+
+    ## floral resource diversity
[…]

### 2018-03-11 analysis/multiSpp-multiSea/src/complete_allInt_filter.R

@@ -1,0 +1,261 @@
+ms.ms.occ <- nimbleCode({
+    ## multi-species priors
+    ## detectablility
+    mu.p.0     ~ dnorm(0,0.01)
+    mu.p.day.1 ~ dnorm(0,0.01)
+    mu.p.day.2 ~ dnorm(0,0.01)
+    sigma.p.0     ~ dunif(0,10)
+    sigma.p.day.1 ~ dunif(0,10)
+    sigma.p.day.2 ~ dunif(0,10)
+
+    ## phi/gam random intercepts
+    mu.phi.0  ~ dnorm(0,0.01)
+    mu.gam.0  ~ dnorm(0,0.01)
+    sigma.phi.0 ~ dunif(0,10)
+    sigma.gam.0 ~ dunif(0,10)
+
+    ## hedgerow area proximity
+    mu.phi.hr.area  ~ dnorm(0,0.01)
+    mu.gam.hr.area  ~ dnorm(0,0.01)
+    sigma.phi.hr.area ~ dunif(0,10)
+    sigma.gam.hr.area ~ dunif(0,10)
+
+    ## semi nat habitat area proximity
+    mu.phi.nat.area  ~ dnorm(0,0.01)
+    mu.gam.nat.area  ~ dnorm(0,0.01)
+    sigma.phi.nat.area ~ dunif(0,10)
+    sigma.gam.nat.area ~ dunif(0,10)
+
+    ## floral resource diversity
+    mu.phi.fra  ~ dnorm(0,0.01)
+    mu.gam.fra  ~ dnorm(0,0.01)
+    sigma.phi.fra ~ dunif(0,10)
+    sigma.gam.fra ~ dunif(0,10)
+
+    ## diet breadth
+    mu.phi.k  ~ dnorm(0,0.01)
+    mu.gam.k  ~ dnorm(0,0.01)
+    sigma.phi.k ~ dunif(0,10)
+    sigma.gam.k ~ dunif(0,10)
+
+    ## body size
+    mu.phi.B  ~ dnorm(0,0.01)
+    mu.gam.B  ~ dnorm(0,0.01)
+    sigma.phi.B ~ dunif(0,10)
+    sigma.gam.B ~ dunif(0,10)
+
+    ## interaction between hedgerow proximity and floral resources
+    ## (habitat quality)
+    mu.phi.hr.area.fra  ~ dnorm(0,0.01)
+    mu.gam.hr.area.fra  ~ dnorm(0,0.01)
[…]

### 2018-03-11 analysis/multiSpp-multiSea/src/initialize.R

@@ -1,0 +1,21 @@
+library(devtools)
+## install_github("nimble-dev/nimble",
+##                ref = "devel",
+##                subdir = "packages/nimble")
+
+library(nimble)
+library(igraph)
+source("../all/plotting.R")
+source("../all/runNimble.R")
+source("../all/comparMCMCs_withMonitors.R")
+source("../all/samplers/sampler_crossLevel_new.R")
+source("../all/samplers/leaveOutSampler.R")
+
+dir.create(file.path("../../../occupancy_saved/saved/multiSpp-multiSea/saved"),
+           showWarnings = FALSE)
+save.dir <-  "../../../occupancy_saved/saved/multiSpp-mutliSea/saved"
+
+## mcmc settings
+scale <- 1e3
+burnin <- 1e2*scale
+niter <- (1e3)*scale

### 2018-03-11 analysis/multiSpp-singleSea/compare.R

@@ -1,16 +1,11 @@
+## setwd('~/Dropbox/occupancy')
 rm(list=ls())
 library(nimble)
 library(coda)
-setwd('~/Dropbox')
-setwd('occupancy/analysis/multiSpp-singleSea')
+setwd('analysis/multiSpp-singleSea')
 source('../all/plotting.R')
 source('src/initialize.R')
-
-thinChains <- function(chains, nthin){
-    apply(chains, c(1,2), function(x){
-        x[!seq(nthin, length(x))] <- NULL
-    })
-}
+source('src/plotPosteriorGroups.R')
 
 ## original model jags and nimble
 load(file=file.path(save.dir, "orig.Rdata"))
@@ -45,8 +40,8 @@ ms.ss.subsamp[[1]] <- rename_MCMC_comparison_method('nimbleSubsamp',
 ## compare mcmcs
 ms.ss.occ.all <- combine_MCMC_comparison_results(ms.ss.orig[[1]],
                                                  ms.ss.filter[[1]],
-                                                 ## ms.ss.crosslevel[[1]],
-                                                 ## ms.ss.subsamp[[1]],
+                                                 ms.ss.crosslevel[[1]],
+                                                 ms.ss.subsamp[[1]],
                                                  name = "ms.ss")
 save(ms.ss.occ.all, file=file.path(save.dir, "combined.Rdata"))
 
@@ -58,61 +53,7 @@ make_MCMC_comparison_pages(ms.ss.occ.all,
 
 checkChains(ms.ss.occ.all$ms.ss$samples,
             f.path = file.path(save.dir,
-
-## paramter groups
-params <- dimnames(ms.ss.occ.all$ms.ss$summary)[[3]]
-groups <- list()
-i <- 1
-while(length(params) > 0){
-    id <- agrep(params[1], params, max.distance = 0.2)
-    groups[[i]] <- params[id]
-    params <- params[-id]
-    i <- i + 1
-}
-
-f <- function(){
-    layout(matrix(1:3, ncol=1))
-    par(oma=c(3,1,1,1))
-    cols <- rainbow(dim(ms.ss.occ.all$ms.ss$summary)[1])
-    for(group in groups){
-        if(length(group) > 1){
-            for(i in 1:dim(ms.ss.occ.all$ms.ss$summary)[1]){
-                this.samp <- ms.ss.occ.all$ms.ss$summary[i,,group]
-                xs <- jitter(1:dim(this.samp)[2])
[…]

### 2018-03-11 analysis/multiSpp-singleSea/filtering.R

@@ -1,6 +1,6 @@
+## setwd('~/Dropbox/occupancy/')
 rm(list=ls())
-setwd('~/Dropbox')
-setwd('occupancy/analysis/multiSpp-singleSea')
+setwd('analysis/multiSpp-singleSea')
 
 source('src/initialize.R')
 ## don't agument data
@@ -12,11 +12,11 @@ model.input <- prepMutiSpData(survey.data,
                               n.zeros,
                               monitors)
 
-load(file=file.path(save.dir, "filter.Rdata"))
-## CHANGE: correctly process intial values, from filtering MCMC output
-## -DT
-model.input$inits <- c(model.input$inits,
-                       genInits(ms.ss.filter[[1]]$summary["nimble", "median",]))
+## load(file=file.path(save.dir, "filter.Rdata"))
+## ## CHANGE: correctly process intial values, from filtering MCMC output
+## ## -DT
+## model.input$inits <- c(model.input$inits,
+##                        genInits(ms.ss.filter[[1]]$summary["nimble", "median",]))
 
 ## *********************************************************************
 ## multi-species site-occupancy models: vectorized with custom
@@ -119,153 +119,153 @@ ms.ss.filter <- compareMCMCs_withMonitors(input1,
 save(ms.ss.filter, file=file.path(save.dir, "filter.Rdata"))
 
 
-## *********************************************************************
-## ## CPPP
-## *********************************************************************
-load(file=file.path(save.dir, "filter.Rdata"))
+## ## *********************************************************************
+## ## ## CPPP
+## ## *********************************************************************
+## load(file=file.path(save.dir, "filter.Rdata"))
 
-occ.R.model <- nimbleModel(code=ms.ss.occ,
-                           constants=input1$constants,
-                           data=input1$data,
-                           inits=input1$inits,
-                           check=FALSE)
+## occ.R.model <- nimbleModel(code=ms.ss.occ,
+##                            constants=input1$constants,
+##                            data=input1$data,
+##                            inits=input1$inits,
+##                            check=FALSE)
 
-occ.mcmc <- buildMCMC(occ.R.model)
-occ.C.model <- compileNimble(occ.R.model)
-occ.C.mcmc <- compileNimble(occ.mcmc, project = occ.R.model)
+## occ.mcmc <- buildMCMC(occ.R.model)
[…]

### 2018-03-11 analysis/multiSpp-singleSea/original.R

@@ -1,6 +1,6 @@
+## setwd('~/Dropbox/occupancy')
 rm(list=ls())
-setwd('~/Dropbox')
-setwd('occupancy/analysis/multiSpp-singleSea')
+setwd('analysis/multiSpp-singleSea')
 
 source('src/initialize.R')
 ## don't agument data
@@ -12,13 +12,6 @@ model.input <- prepMutiSpData(survey.data,
                               n.zeros,
                               monitors,
                               remove.zs=FALSE)
-
-load(file=file.path(save.dir, "filter.Rdata"))
-## CHANGE: correctly process intial values, from filtering MCMC output
-## -DT
-model.input$inits <- c(model.input$inits,
-                       genInits(ms.ss.filter[[1]]$summary["nimble", "median",]))
-rm(ms.ss.filter)
 
 ## *********************************************************************
 ## multi-species site-occupancy models: original
@@ -78,6 +71,7 @@ ms.ss.occ <- nimbleCode({
         b2[i] ~ dnorm(mu.b2, tau.b2)
         ## Create a loop to estimate the Z matrix (true occurrence for
         ## species i at point j).
+
         for (j in 1:num.points) {
             logit(psi[j,i]) <- u.cato[i]*(1-habitat.ind[j]) +
                 u.fcw[i]*habitat.ind[j] +
@@ -100,8 +94,6 @@ ms.ss.occ <- nimbleCode({
         }
     }
 })
-
-
 
 input1 <- c(code=ms.ss.occ, model.input)
 

### 2018-03-11 analysis/multiSpp-singleSea/original_nimble.R

@@ -1,126 +1,0 @@
-rm(list=ls())
-setwd('occupancy/analysis/multiSpp-singleSea')
-
-## Value for data augmentation (n_zeroes is an arbitrarily large value)
-n_zeroes <- 50
-source('src/initialize.R')
-
-## *********************************************************************
-## multi-species site-occupancy models
-## *********************************************************************
-## Since calculations with date_linear and date_quadratic are now
-## vectorized, we'll set the NAs to 0
-model_data$date_linear[is.na(model_data$date_linear)] <- 0
-model_data$date_quadratic[is.na(model_data$date_quadratic)] <- 0
-
-multispecies_occupancy <- nimbleCode({
-    ## Define prior distributions for community-level model parameters
-    omega ~ dunif(0,1)
-
-    cato_occ_mean ~ dunif(0,1)
-    mu_ucato <- log(cato_occ_mean) - log(1-cato_occ_mean)
-
-    fcw_occ_mean ~ dunif(0,1)
-    mu_ufcw <- log(fcw_occ_mean) - log(1-fcw_occ_mean)
-
-    cato_det_mean ~ dunif(0,1)
-    mu_vcato <- log(cato_det_mean) - log(1-cato_det_mean)
-
-    fcw_det_mean ~ dunif(0,1)
-    mu_vfcw <- log(fcw_det_mean) - log(1-fcw_det_mean)
-
-    ## random effects
-    sigma_ucato ~ dunif(0, 100)
-    sigma_ufcw ~ dunif(0, 100)
-    mu_a1 ~ dnorm(0, 0.001)
-    sigma_a1 ~ dunif(0, 100)
-    mu_a2 ~ dnorm(0, 0.001)
-    sigma_a2 ~ dunif(0, 100)
-    mu_a3 ~ dnorm(0, 0.001)
-    sigma_a3 ~ dunif(0, 100)
-    mu_a4 ~ dnorm(0, 0.001)
-    sigma_a4 ~ dunif(0, 100)
-
-    sigma_vcato ~ dunif(0, 100)
-    sigma_vfcw ~ dunif(0, 100)
-    mu_b1 ~ dnorm(0, 0.001)
-    sigma_b1 ~ dunif(0, 100)
-    mu_b2 ~ dnorm(0, 0.001)
-    sigma_b2 ~ dgamma(0.1,0.1)
-
-
-    for (i in 1:(num_species + n_zeroes)) {
[…]

### 2018-03-11 analysis/multiSpp-singleSea/src/initialize.R

@@ -34,5 +34,7 @@ monitors <- c('mu.a1','mu.a2','mu.a3','mu.a4',
               'cato.det.mean', 'fcw.det.mean',
               'sigma.ucato', 'sigma.vcato', 'sigma.ufcw',
               'sigma.vfcw', 'mu.b1', 'mu.b2', 'sigma.b1',
-              'sigma.b2', 'u.cato', 'u.fcw',
-              'a1', 'a2', 'a3', 'a4', 'v.cato', 'v.fcw', 'b1', 'b2')
+              'sigma.b2')
+## mid level params kept track of for diagnosing problems with model
+## 'u.cato', 'u.fcw',
+## 'a1', 'a2', 'a3', 'a4', 'v.cato', 'v.fcw', 'b1', 'b2')

### 2018-03-11 analysis/multiSpp-singleSea/src/plotPosteriorGroups.R

@@ -1,0 +1,41 @@
+
+plotPosteriorGroups <- function(){
+    layout(matrix(1:3, ncol=1))
+    par(oma=c(3,1,1,1))
+    cols <- rainbow(dim(ms.ss.occ.all$ms.ss$summary)[1])
+    for(group in groups){
+        if(length(group) > 1){
+            for(i in 1:dim(ms.ss.occ.all$ms.ss$summary)[1]){
+                this.samp <- ms.ss.occ.all$ms.ss$summary[i,,group]
+                xs <- jitter(1:dim(this.samp)[2])
+                if(i == 1){
+                    plot(x=xs, y=this.samp["mean",], pch=16,
+                         col=cols[i],
+                         ylim=range(c(ms.ss.occ.all$ms.ss$summary[,'CI95_upp',
+                                                                  group],
+                                      ms.ss.occ.all$ms.ss$summary[,'CI95_low',
+                                                                  group])),
+                         xaxt="n",
+                         ylab="Estimate",
+                         xlab="")
+
+                    axis(1, at=1:dim(this.samp)[2],
+                         labels=FALSE)
+                    text(x=1:dim(this.samp)[2],
+                         y=par()$usr[3]-0.1*(par()$usr[4]-par()$usr[3]),
+                         labels=group, srt=45, adj=1, xpd=TRUE)
+                } else{
+                    points(x=xs,
+                           y=this.samp["mean",], pch=16, col=cols[i])
+                }
+                arrows(y1=this.samp['CI95_upp',],
+                       y0=this.samp['CI95_low',],
+                       x0=xs,
+                       code=0, angle=90, length=0.02, lwd=1, col=cols[i])
+            }
+        }
+
+        legend("topright", legend=dimnames(ms.ss.occ.all$ms.ss$summary)[[1]],
+               pch=16, col=cols)
+    }
+}

### 2018-03-11 analysis/multiSpp-singleSea/src/reformatData.R

@@ -198,17 +198,17 @@ prepMutiSpData <- function(survey.data,
     inits <-list(Z=zinits,
                  omega = omega.draw,
                  w = c(rep(1, num.species),
-                       rbinom(n.zeroes, size = 1, prob = omega.draw)))
-                 ## u.cato = rnorm(num.species + n.zeroes),
-                 ## v.cato = rnorm(num.species + n.zeroes),
-                 ## u.fcw = rnorm(num.species + n.zeroes) ,
-                 ## v.fcw = rnorm(num.species + n.zeroes),
-                 ## a1 = rnorm(num.species + n.zeroes),
-                 ## a2 = rnorm(num.species + n.zeroes),
-                 ## a3 = rnorm(num.species + n.zeroes),
-                 ## a4 = rnorm(num.species + n.zeroes),
-                 ## b1 = rnorm(num.species + n.zeroes),
-                 ## b2 = rnorm(num.species + n.zeroes))
+                       rbinom(n.zeroes, size = 1, prob = omega.draw)),
+                 u.cato = rnorm(num.species + n.zeroes),
+                 v.cato = rnorm(num.species + n.zeroes),
+                 u.fcw = rnorm(num.species + n.zeroes) ,
+                 v.fcw = rnorm(num.species + n.zeroes),
+                 a1 = rnorm(num.species + n.zeroes),
+                 a2 = rnorm(num.species + n.zeroes),
+                 a3 = rnorm(num.species + n.zeroes),
+                 a4 = rnorm(num.species + n.zeroes),
+                 b1 = rnorm(num.species + n.zeroes),
+                 b2 = rnorm(num.species + n.zeroes))
 
     ## constants
     constants <- list(num.species = num.species,

### 2018-03-11 analysis/singleSpp-multiSea/compare.R

@@ -1,7 +1,9 @@
 rm(list=ls())
 library(nimble)
 library(coda)
-setwd("occupancy/analysis/singleSpp-multiSea")
+setwd("~/Dropbox/occupancy")
+setwd("analysis/singleSpp-multiSea")
+
 source('../all/plotting.R')
 source('src/initialize.R')
 
@@ -69,19 +71,22 @@ plotEffSize(by.config, by.param, f.path= file.path(save.dir,
                                 "SingleSpp-MultiSea",
             at=50, adj1=1, adj2=0.3, widths=c(4.5, 8.5))
 
-
+## difference between default Jags and Nimble
 (ss.ms.occ.all$ss.ms$efficiency$mean["JAGS-latent"] -
  ss.ms.occ.all$ss.ms$efficiency$mean["NIMBLE-latent"])/
     max(ss.ms.occ.all$ss.ms$efficiency$mean["JAGS-latent"])
 
+## jags and filtering
 (ss.ms.occ.all$ss.ms$efficiency$mean["JAGS-latent"] -
  ss.ms.occ.all$ss.ms$efficiency$mean["NIMBLE-filter"])/
     max(ss.ms.occ.all$ss.ms$efficiency$mean["JAGS-latent"])
 
+## jags and subsampling
 (ss.ms.occ.all$ss.ms$efficiency$mean["JAGS-latent"] -
  ss.ms.occ.all$ss.ms$efficiency$mean["NIMBLE-subsample"])/
     max(ss.ms.occ.all$ss.ms$efficiency$mean["JAGS-latent"])
 
+## subsampling and nimble defaults
 (ss.ms.occ.all$ss.ms$efficiency$mean["NIMBLE-subsample"] -
  ss.ms.occ.all$ss.ms$efficiency$mean["NIMBLE-latent"])/
     max(ss.ms.occ.all$ss.ms$efficiency$mean["NIMBLE-subsample"])

### 2018-03-11 analysis/singleSpp-multiSea/filtering.R

@@ -1,6 +1,6 @@
 rm(list=ls())
-
-setwd("occupancy/analysis/singleSpp-multiSea")
+## setwd("~/Dropbox/occupancy/")
+setwd("analysis/singleSpp-multiSea")
 source('src/initialize.R')
 set.seed(444)
 data <- genDynamicOccData()

### 2018-03-11 analysis/singleSpp-multiSea/src/dynamicOcc.R

@@ -112,8 +112,6 @@ registerDistributions(list(dbern_matrix = list(
 ## DynamicOccupancy removes the z's and muZ's from the model and computes
 ## the probability of all reps over all years for one site.
 dDynamicOccupancy <- nimbleFunction(
-    ## I've checked that this runs and compiles, but I haven't tested if
-    ## I got the logic right!
     run = function(x = double(2),
                    nrep = double(),
                    psi1 = double(),
@@ -121,7 +119,7 @@ dDynamicOccupancy <- nimbleFunction(
                    gamma = double(1),
                    p = double(1), # should double(2)
                    log = double(0, default = 0)) {
-        prob1 <- psi1 * p[1]
+
         numObs <- sum(x[,1]) ## do I have the right orientation?
         ## prob of the occupied sites out of the total sites given p
         ProbOccAndCount <- psi1 * dbinom(numObs, size = nrep, p = p[1], log = 0)

### 2018-03-11 analysis/singleSpp-multiSea/src/initialize.R

@@ -3,6 +3,13 @@ library(devtools)
 ##                ref = "devel",
 ##                subdir = "packages/nimble")
 
+
+install_github("nimble-dev/nimble",
+               ref = "faster-vector-passing",
+               subdir = "packages/nimble")
+
+
+
 library(nimble)
 library(igraph)
 source("src/dynamicOcc.R")
